## 📦 Build the phyloseq object (18S)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)

# ---- paths ----
datadir <- "data/18S"

# ---- load DADA2 outputs ----
seqtab.nochim <- readRDS(file.path(datadir, "seqtab_nochim.rds"))
taxa          <- readRDS(file.path(datadir, "taxa.rds"))

# Make sure taxonomy is a proper matrix and has consistent column names
taxa <- as.matrix(taxa)
target_cols <- c(
  "Kingdom","Subkingdom_1","Subkingdom_2","Phylum","Subphylum","Subphylum_1","Superclass",
  "Class","Subclass","Infraclass","Superorder","Order","Suborder_1","Suborder_2","Suborder_3",
  "Family","Subfamily","Genus","Subgenus","Species"
)
colnames(taxa) <- target_cols[seq_len(ncol(taxa))]

# Load metadata
metadat <- readRDS(file.path(datadir, "metadata_18S.rds")) 

# Create a Phyloseq object

rownames(seqtab.nochim) <- sub("_$", "", rownames(seqtab.nochim))

ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = F), 
               sample_data(metadat), 
               tax_table(taxa))
cat("🔹 ** A summary of the phyloseq object**\n")
ps

# storing the ASV sequences and assigning them with a name and then renaming the taxa object
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))

```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Print sample names, taxonomy ranks, and metadata fields — with explanations
cat("🔹 **These are the metadata variables:**\n")
print(sample_variables(ps))
```
## 📊 Data visualisation

In this section, we explore the RoCSI-CPR 18S community structure using the
`phyloseq` object (`ps`) generated above. We start by visualising read depth
per sample, followed by basic taxonomic composition summaries.

### Read depth per sample
```{r, echo=FALSE, message=FALSE, warning=FALSE}
read_depth <- data.frame(
Sample = sample_names(ps),
Reads = sample_sums(ps),
check.names = FALSE
)

ggplot(read_depth, aes(x = Sample, y = Reads)) +
geom_col() +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
scale_x_discrete(
    labels = function(x) substr(x, 1, 6), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  ) +
labs(
title = "Sequencing depth per sample",
x = "Sample",
y = "Number of non-chimeric reads"
)
```

### Alpha diversity

```{r, echo=FALSE, message=FALSE, warning=FALSE}
sample_data(ps)$Water_masses <- ifelse(sample_data(ps)$lon < -10,
                                       "Open Ocean", "Shelf")

# Make sure Water_masses is a factor (optional, but useful for consistent plotting)
sample_data(ps)$Water_masses <- factor(sample_data(ps)$Water_masses, 
                                       levels = c("Open Ocean", "Shelf"))

plot_richness(ps, x = "lon", measures = c("Shannon", "Observed"), color = "Water_masses", 
              title = "RoCSI (18S, MZGdb)")
```
::: {.callout-note}
### 💬 Discussion — Alpha diversity patterns

Both richness (Observed ASVs) and Shannon diversity differ between water masses, although the contrasts are not very large.

The Open Ocean samples exhibit higher richness and slightly higher evenness, likely reflecting a more stable and stratified offshore environment that supports a broader range of protist niches.

Shelf samples have lower richness and evenness, consistent with a more turbulent, nutrient-rich system dominated by a few fast-growing taxa.  
:::

### Ordination

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ord <- ordinate(ps, method = "PCoA", distance = "bray")

# Create the ordination plot with ellipses
plot_ordination(ps, ord, color = "Water_masses") +
  geom_point(size = 3) +
  ggtitle("Ordination – RoCSI (18S, MZGdb)") +
  stat_ellipse(aes(group = Water_masses), type = "norm", linetype = "dashed", size = 1) +
  scale_color_manual(values = c("Open Ocean" = "#377EB8", "Shelf" = "#FF6666")) +
  theme_minimal()
```
::: {.callout-note}
### 💬 Discussion — Community Differences Between Water Masses

The ordination shows clear but not absolute separation between Shelf and Open Ocean communities. Offshore samples cluster tightly, indicating more homogeneous conditions, while shelf samples are more dispersed, reflecting greater environmental variability. Overall, the pattern suggests that water-mass structure shapes protist community composition along the RoCSI–CPR transect.  
:::

### Phylum composition

```{r, echo=FALSE, message=FALSE, warning=FALSE}
my_colors <- c("#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A",  
               "#984EA3", "#A6D854", "#FFFF33", "#FF7F00", "#F781BF", "#999999",                  "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#FFD92F", 
               "#E5C494", "#B3B3B3", "#A65628","#FFFFB3", "#BEBADA", 
               "#FB8072", "#80B1D3", "#FDB462")

classGlommed = tax_glom(ps, "Phylum")

# transform to proportions
ps1_rel <- transform_sample_counts(classGlommed, function(x) x / sum(x))
# Ensure 'lon' is a factor
ps1_rel@sam_data$lon <- as.factor(ps1_rel@sam_data$lon)

plot_bar(ps1_rel, fill = "Phylum", x = "lon") + 
  scale_fill_manual(values = my_colors) + theme_classic() +
  geom_bar(stat = "identity", width = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Relative abundance (%)", x = "Longitude") +
  ggtitle("Protist Phyla – RoCSI (18S, MZGdb)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
  scale_x_discrete(
    labels = function(x) substr(x, 1, 5), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  )
```

::: {.callout-note}
### 💬 Discussion — Protist Community Structure

The 18S protist communities along the RoCSI transect are dominated by **Cercozoans** (Heterotrophic protists/grazers) and **Myzozoans** (subphylum Dinoflagellata). **Radiolarians** (unicellular eukaryotes) are abundant in the open-water section of the transect, while **Bacillariophyta** (Diatoms) are abundant at the beginning of the cruise.

Is this phyla composition ecologically coherent with expected shelf/open-ocean plankton dynamics during early spring in the English channel/Celtic Sea?
:::

### Top 10 genera 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
### --------------- Using the microbiome package ----------------
#BiocManager::install("microbiome")
library(microbiome)

# Collapse to genus level
ps_genus <- tax_glom(ps, taxrank = "Genus")

# Transform to relative abundance
ps_rel <- transform(ps_genus, "compositional")

# Get top 10 genera by mean relative abundance
top10 <- top_taxa(ps_rel, n = 10)  # this returns taxa_names, not Genus strings

# Melt *full* ps_rel, not pruned
df <- psmelt(ps_rel)
df$lon <- as.factor(df$lon)

# Create a new genus column where everything not in top10 becomes "Other"
df <- df %>%
  mutate(
    Genus_plot = as.character(Genus),
    Genus_plot = if_else(OTU %in% top10, Genus_plot, "Other")
  ) %>%
  group_by(Sample, lon, Genus_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Optional: put "Other" at the end of the legend
df$Genus_plot <- factor(
  df$Genus_plot,
  levels = c(setdiff(sort(unique(df$Genus_plot)), "Other"), "Other")
)

# Pick a clear neutral colour for "Other"
other_color <- "#C0C0C0"

# Fit your colours + "Other" at the end
plot_colors <- c(my_colors[1:10], other_color)

# Plot
ggplot(df, aes(x = lon, y = Abundance, fill = Genus_plot)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = plot_colors) + 
  labs(
    title = "Top 10 Genera + Other",
    x = "Longitude",
    y = "Relative Abundance"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(
    labels = function(x) substr(x, 1, 5),
    breaks  = function(x) x[seq(1, length(x), by = 4)]
  )

```
::: {.callout-note}
### 💬 Discussion — Dominant protist genera

Breaking down the 18S community at the genus level reveals a mixture of **picophytoplankton (Bathycoccus), mixotrophic and autotrophic dinoflagellates (Karlodinium, Karenia, Prorocentrum and Lepidodinium), and heterotrophic grazers (Goniomonas, Protoperidinium)**.

The high abundance of dinoflagellates **Karlodinium, Karenia and Lepidodinium** at the transition between shelp to open-ocean might be linked to the observed phytoplankton blooms seen from the satelite data and chlorophll index of the CPR (see below).

Let's note the high abundance of **Goniomonas** (Phagotrophic micrograzer) in the open-ocean!

Probably: shelf --> well mixed, Open-ocean --> more stratified...

**Note**: The "other" genius (including all the less abundant genera) is large, especially at low longitudes...
:::
