# 18S Dataset

## 📦 Build the phyloseq object (18S)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)

# ---- paths ----
datadir <- "data/18S"

# ---- load DADA2 outputs ----
seqtab.nochim <- readRDS(file.path(datadir, "seqtab_nochim.rds"))
taxa          <- readRDS(file.path(datadir, "taxa.rds"))

# Make sure taxonomy is a proper matrix and has consistent column names
taxa <- as.matrix(taxa)
target_cols <- c(
  "Kingdom","Subkingdom_1","Subkingdom_2","Phylum","Subphylum","Subphylum_1","Superclass",
  "Class","Subclass","Infraclass","Superorder","Order","Suborder_1","Suborder_2","Suborder_3",
  "Family","Subfamily","Genus","Subgenus","Species"
)
colnames(taxa) <- target_cols[seq_len(ncol(taxa))]

# Load metadata
metadat <- readRDS(file.path(datadir, "metadata_18S.rds")) 

# Create a Phyloseq object

rownames(seqtab.nochim) <- sub("_$", "", rownames(seqtab.nochim))

ps_18s <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = F), 
               sample_data(metadat), 
               tax_table(taxa))
saveRDS(ps_18s, "data/ps_18s.rds")
  
cat("🔹 A summary of the phyloseq object \n")
ps_18s

# storing the ASV sequences and assigning them with a name and then renaming the taxa object
dna <- Biostrings::DNAStringSet(taxa_names(ps_18s))
names(dna) <- taxa_names(ps_18s)
ps_18s <- merge_phyloseq(ps_18s, dna)
taxa_names(ps_18s) <- paste0("ASV", seq(ntaxa(ps_18s)))

```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Print sample names, taxonomy ranks, and metadata fields — with explanations
cat("🔹 These are the metadata variables:\n")
print(sample_variables(ps_18s))
```

## 📊 Data visualisation

In this section, we explore the community structure of the RoCSI 18S dataset using the
`phyloseq` object (`ps_18s`) generated above. We start by visualising read depth
per sample, followed by basic taxonomic composition summaries.

### Read depth per sample
```{r, echo=FALSE, message=FALSE, warning=FALSE}
read_depth <- data.frame(
Sample = sample_names(ps_18s),
Reads = sample_sums(ps_18s),
check.names = FALSE
)

ggplot(read_depth, aes(x = Sample, y = Reads)) +
geom_col() +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
scale_x_discrete(
    labels = function(x) substr(x, 1, 6), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  ) +
labs(
title = "Sequencing depth per sample",
x = "Sample",
y = "Number of non-chimeric reads"
)
```

### Alpha diversity

```{r, echo=FALSE, message=FALSE, warning=FALSE}
sample_data(ps_18s)$Water_masses <- ifelse(sample_data(ps_18s)$lon < -10,
                                       "Open Ocean", "Shelf")

# Make sure Water_masses is a factor (optional, but useful for consistent plotting)
sample_data(ps_18s)$Water_masses <- factor(sample_data(ps_18s)$Water_masses, levels = c("Open Ocean", "Shelf"))

plot_richness(
  ps_18s,
  x = "lon",
  measures = c("Observed", "Shannon", "Simpson"),
  color = "Water_masses",
  title = "RoCSI (18S, MZGdb)"
) +
  scale_color_manual(
    values = c(
      "Open Ocean" = "#2E86C1",
      "Shelf" = "#C0392B"
    )
  )

```
::: {.callout-note}
### 💬 Discussion — Alpha diversity patterns

Both Observed richness (number of ASVs) and Simpson diversity differ between water masses, although the contrasts are not very large.

The Open Ocean samples exhibit higher richness and slightly higher evenness, likely reflecting a more stable and stratified offshore environment that supports a broader range of protist niches.

Shelf samples have lower richness and evenness, consistent with a more turbulent, nutrient-rich system dominated by a few fast-growing taxa.  
:::

### Ordination

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ord <- ordinate(ps_18s, method = "PCoA", distance = "bray")

# Create the ordination plot with ellipses
plot_ordination(ps_18s, ord, color = "Water_masses") +
  geom_point(size = 3) +
  ggtitle("Ordination – RoCSI (18S, MZGdb)") +
  stat_ellipse(aes(group = Water_masses), type = "norm", linetype = "dashed", size = 1) +
  scale_color_manual(values = c("Open Ocean" = "#2E86C1","Shelf" = "#C0392B")) +
  theme_minimal()
```
::: {.callout-note}
### 💬 Discussion — Community Differences Between Water Masses

The ordination shows clear but not absolute separation between Shelf and Open Ocean communities. Offshore samples cluster tightly, indicating more homogeneous conditions, while shelf samples are more dispersed, reflecting greater environmental variability. Overall, the pattern suggests that water-mass structure shapes protist community composition along the RoCSI–CPR transect.

Although Open Ocean samples show slightly higher within-sample diversity (richness, Shannon), the ordination reveals that Shelf samples are more heterogeneous in composition.
This indicates that while offshore communities are individually more diverse and compositionally similar, shelf communities are shaped by more variable environmental conditions, leading to greater between-sample variation.

**NOTE**: the 5 samples clustered together in the lower part of the plot are the first 5 samples taken in first transect (close to Southampton). These samples occupy an extreme position in ordination space, suggesting strong compositional differences, which may be associated with dominance effects.
:::

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Extract sample coordinates from the PCoA result
site <- as.data.frame(ord$vectors)
site$SampleID <- rownames(site)

# Get the 5 lowest values on Axis.2 (bottom of plot)
bottom5 <- site[order(site$Axis.2), ][1:5, c("SampleID", "Axis.1", "Axis.2")]
bottom5
```

### Phylum composition

```{r, echo=FALSE, message=FALSE, warning=FALSE}
my_colors <- c("#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A",  
               "#984EA3", "#A6D854", "#FFFF33", "#FF7F00", "#F781BF", "#999999",                  "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#FFD92F", 
               "#E5C494", "#B3B3B3", "#A65628","#FFFFB3", "#BEBADA", 
               "#FB8072", "#80B1D3", "#FDB462")

classGlommed = tax_glom(ps_18s, "Phylum")

# transform to proportions
ps_18s_rel <- transform_sample_counts(classGlommed, function(x) x / sum(x))
# Ensure 'lon' is a factor
ps_18s_rel@sam_data$lon <- as.factor(ps_18s_rel@sam_data$lon)

plot_bar(ps_18s_rel, fill = "Phylum", x = "lon") + 
  scale_fill_manual(values = my_colors) + theme_classic() +
  geom_bar(stat = "identity", width = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Relative abundance (%)", x = "Longitude") +
  ggtitle("Protist Phyla – RoCSI (18S, MZGdb)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
  scale_x_discrete(
    labels = function(x) substr(x, 1, 5), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  )
```

::: {.callout-note}
### 💬 Discussion — Protist Community Structure

The 18S protist communities along the RoCSI transect are dominated by **Cercozoans** (Heterotrophic protists/grazers) and **Myzozoans** (subphylum Dinoflagellata). **Radiolarians** (unicellular eukaryotes) are abundant in the open-water section of the transect, while **Bacillariophyta** (Diatoms) are abundant at the beginning of the cruise.

Is this phyla composition ecologically coherent with expected shelf/open-ocean plankton dynamics during early spring in the English channel/Celtic Sea?
:::

### Phylum composition - including NAs

```{r,echo=FALSE, message=FALSE, warning=FALSE}
classGlommed = tax_glom(ps_18s, "Phylum", NArm = FALSE)

# transform to proportions
ps_18s_rel <- transform_sample_counts(classGlommed, function(x) x / sum(x))
# Ensure 'lon' is a factor
ps_18s_rel@sam_data$lon <- as.factor(ps_18s_rel@sam_data$lon)

plot_bar(ps_18s_rel, fill = "Phylum", x = "lon") + 
  scale_fill_manual(values = my_colors) + theme_classic() +
  geom_bar(stat = "identity", width = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Relative abundance (%)", x = "Longitude") +
  ggtitle("Protist Phyla – RoCSI (18S, MZGdb) - including NAs")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
  scale_x_discrete(
    labels = function(x) substr(x, 1, 5), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  )
```

### Top 10 genera 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
### --------------- Using the microbiome package ----------------
#BiocManager::install("microbiome")
library(microbiome)

# Collapse to genus level
ps_genus <- tax_glom(ps_18s, taxrank = "Genus")

# Transform to relative abundance
ps_rel <- transform(ps_genus, "compositional")

# Get top 10 genera by mean relative abundance
top10 <- top_taxa(ps_rel, n = 10)  # this returns taxa_names, not Genus strings

# Melt *full* ps_rel, not pruned
df <- psmelt(ps_rel)
df$lon <- as.factor(df$lon)

# Create a new genus column where everything not in top10 becomes "Other"
df <- df %>%
  mutate(
    Genus_plot = as.character(Genus),
    Genus_plot = if_else(OTU %in% top10, Genus_plot, "Other")
  ) %>%
  group_by(Sample, lon, Genus_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Optional: put "Other" at the end of the legend
df$Genus_plot <- factor(
  df$Genus_plot,
  levels = c(setdiff(sort(unique(df$Genus_plot)), "Other"), "Other")
)

# Pick a clear neutral colour for "Other"
other_color <- "#C0C0C0"

# Fit your colours + "Other" at the end
plot_colors <- c(my_colors[1:10], other_color)

# Plot
ggplot(df, aes(x = lon, y = Abundance, fill = Genus_plot)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = plot_colors) + 
  labs(
    title = "Top 10 Genera + Other",
    x = "Longitude",
    y = "Relative Abundance"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(
    labels = function(x) substr(x, 1, 5),
    breaks  = function(x) x[seq(1, length(x), by = 4)]
  )

```
::: {.callout-note}
### 💬 Discussion — Dominant protist genera

Breaking down the 18S community at the genus level reveals a mixture of **picophytoplankton (Bathycoccus), mixotrophic and autotrophic dinoflagellates (Karlodinium, Karenia, Prorocentrum and Lepidodinium), and heterotrophic grazers (Goniomonas, Protoperidinium)**.

The high abundance of dinoflagellates **Karlodinium, Karenia and Lepidodinium** at the transition between shelp to open-ocean might be linked to the observed phytoplankton blooms seen from the satelite data and chlorophll index of the CPR (see below).

Let's note the high abundance of **Goniomonas** (Phagotrophic micrograzer) in the open-ocean!

Probably: shelf --> well mixed, Open-ocean --> more stratified...

**Note**: The "other" genius (including all the less abundant genera) is large, especially at low longitudes...
**IMPORTANT**: "other" also includes **NAs** which can be not-assigned ASVs such as zooplankton. For example, ASV2 is the most abundant ASV in 01-CPR_1_ID_1 but has only Kingdom = Chromista as taxonomy. BLasting the sequence shows that it is the 18S sequence of **Centropages sp.**
ASV3 = Oithona sp.
:::

### Main Phytoplankton genera

```{r,echo=FALSE, message=FALSE, warning=FALSE}
## 1) Keep only phytoplankton phyla
phyto_phyla <- c(
  "Bacillariophyta",
  "Haptophyta",
  "Chlorophyta",
  "Cryptophyta",
  "Rhodophyta",
  "Ochrophyta",
  "Myzozoa"
)

ps_phyto <- subset_taxa(ps_18s, Phylum %in% phyto_phyla)
ps_phyto <- prune_taxa(taxa_sums(ps_phyto) > 0, ps_phyto)

## 2) Collapse to genus + drop unassigned genera
ps_genus <- tax_glom(ps_phyto, taxrank = "Genus", NArm = FALSE)
ps_genus <- subset_taxa(ps_genus, !is.na(Genus) & Genus != "" & Genus != "NA")
ps_genus <- prune_taxa(taxa_sums(ps_genus) > 0, ps_genus)

## 3) Relative abundance (within phytoplankton genera)
ps_rel <- transform(ps_genus, "compositional")

## 4) Keep only top 10 phytoplankton genera
top10 <- top_taxa(ps_rel, n = 10)
ps_top10 <- prune_taxa(top10, ps_rel)

## OPTIONAL: re-normalise so top10 sum to 1
ps_top10 <- transform(ps_top10, "compositional")

## 5) Melt for plotting
df <- psmelt(ps_top10)
df$lon <- as.factor(df$lon)

## Order genera by mean abundance
gen_order <- df %>%
  group_by(Genus) %>%
  summarise(m = mean(Abundance), .groups = "drop") %>%
  arrange(desc(m)) %>%
  pull(Genus)

df$Genus <- factor(df$Genus, levels = gen_order)

## 6) Match colours to genera (IMPORTANT)
plot_colors <- my_colors[1:length(gen_order)]
names(plot_colors) <- gen_order

## 7) Plot
ggplot(df, aes(x = lon, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = plot_colors) +
  labs(
    title = "Top 10 phytoplankton genera",
    x = "Longitude",
    y = "Relative Abundance",
    fill = "Genus"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(
    labels = function(x) substr(x, 1, 5),
    breaks  = function(x) x[seq(1, length(x), by = 4)]
  )
```
Surprisingly the CPR dataset has many occurences of Emiliania huxleyi but none in the 18S dataset...
Let's look for Gephyrocapsa sp. (other accepted genus)

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ps_geo <- subset_taxa(ps_18s, Genus == "Gephyrocapsa")
ps_geo <- prune_taxa(taxa_sums(ps_geo) > 0, ps_geo)

df <- psmelt(ps_geo)
df$lon <- as.numeric(as.character(df$lon))  # if lon is factor

df_sum <- df %>%
  group_by(Sample, lon) %>%
  summarise(Count = sum(Abundance), .groups = "drop")

ggplot(df_sum, aes(x = lon, y = Count)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "Gephyrocapsa read counts along transect",
    x = "Longitude",
    y = "Read count"
  ) +
  theme_minimal()




```

