---
title: "Analysis code for the RoCSI-CPR study"
author: "Nathan Hubot, Susan Evans, Julie Robidart"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
editor: source
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

# Overview

This Quarto document contains the code used to process and compare the RoCSI eDNA metabarcoding dataset with Continuous Plankton Recorder (CPR) microscopy data.

Two metabarcoding markers were analysed:

- **18S rRNA V9**: primarily targeting phytoplankton and other protists.
- **COI mitochondrial marker**: primarily targeting metazoan plankton.

The eDNA and CPR datasets were compared using taxonomic overlap, prevalence across longitudinal sections, and community-level ordination.

# Sequencing and bioinformatic workflow

## Sequencing summary

| Feature | **18S rRNA V9** | **COI** |
|---|---|---|
| **SSP ID** | [SSP202877](https://cgr.liv.ac.uk/illum/SSP202877_b388934ddf413642/) | [SSP200993](https://cgr.liv.ac.uk/illum/SSP200993_bbba6061877a5e12) |
| **Purchase order** | 203631529 | P10817-5 |
| **Sequencing platform** | Illumina MiSeq v2 (2 × 150 bp) | Illumina MiSeq v2 (2 × 250 bp) |
| **Target region** | 18S V9, 1389F–1510R | COI Leray fragment, m1COIintF–jgHCO2198 |
| **Expected amplicon size** | ~121 bp | ~313 bp |
| **Raw read depth, mean** | ~150,000 reads | ~200,000 reads |
| **Pre-processing** | Cutadapt v4.5 | Cutadapt v1.2.1 + Sickle v1.200 |
| **Downstream workflow** | Cutadapt → DADA2 → MZG 18S taxonomy | Cutadapt → DADA2 → MZG COI taxonomy |

## Primer sets

### 18S V9

- **Forward**: 5′ ACACTCTTTCCCTACACGACGCTCTTCCGATCTNNNNN **TTGTACACACCGCCC** 3′
- **Reverse**: 5′ GTGACTGGAGTTCAGACGTGTGCTCTTCCGATCT **CCTTCYGCAGGTTCACCTAC** 3′

### COI

- **Forward**: 5′ ACACTCTTTCCCTACACGACGCTCTTCCGATCTNNNNN **GGWACWGGWTGAACWGTWTAYCCYCC** 3′
- **Reverse**: 5′ GTGACTGGAGTTCAGACGTGTGCTCTTCCGATCT **TAIACYTCIGGRTGICCRAARAAYCA** 3′

## Primer removal with Cutadapt

| Parameter | **18S** | **COI** |
|---|---|---|
| Forward primer (`-g`) | TTGTACACACCGCCC | GGWACWGGWTGAACWGTWTAYCCYCC |
| Reverse primer (`-G`) | CCTTCYGCAGGTTCACCTAC | TANACYTCNGGRTGNCCRAARAAYCA |
| `--match-read-wildcards` | yes | yes |
| Minimum overlap | 10 | 20 |
| Max error rate (`-e`) | 0.15 | 0.20 |
| `--discard-untrimmed` | yes | yes |
| `--minimum-length` | 80 bp | 200 bp |

Both datasets were processed with a marker-specific version of the following command:

```bash
cutadapt \
  -g <forward_primer> \
  -G <reverse_primer> \
  --match-read-wildcards \
  --overlap <minimum_overlap> \
  -e <maximum_error_rate> \
  --pair-filter=both \
  --discard-untrimmed \
  --cores=0 \
  -o <trimmed_R1.fastq.gz> \
  -p <trimmed_R2.fastq.gz> \
  <input_R1.fastq.gz> \
  <input_R2.fastq.gz>
```

## DADA2 processing

| Parameter | **18S** | **COI** |
|---|---:|---:|
| `truncLen` | `c(130, 120)` | `c(210, 210)` |
| `maxEE` | `c(2, 3)` | `c(2, 3)` |
| `minLen` | 80 | 200 |
| `minOverlap` in `mergePairs()` | 30 | 90 |
| Reference database | MZG 18S | MZG COI |

::: {.callout-note title="Reference databases"}
Taxonomy was assigned with DADA2 using MetaZooGene reference databases:

- **18S**: `MZGdada2-18s__T2000000__o00__A.fastq`, all microbes and protists, mode A.
- **COI**: `MZGdada2-coi__T4000000__o00__A.fastq`, all invertebrates, mode A.
:::

Both markers were processed with the standard DADA2 workflow: filtering, error learning, dereplication, ASV inference, pair merging, chimera removal, and taxonomic assignment.

# Setup

```{r setup}
library(tidyverse)
library(readxl)
library(phyloseq)
library(vegan)
library(Biostrings)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
library(ggVennDiagram)
library(scales)
library(ggOceanMaps)
library(ggspatial)
library(ggnewscale)
library(cowplot)

set.seed(123)

theme_set(theme_minimal(base_size = 12))
```

# Helper functions

```{r helper-functions}
# Taxonomic ranks returned by the MZG DADA2 databases.
mzg_ranks <- c(
  "Kingdom", "Subkingdom_1", "Subkingdom_2", "Phylum", "Subphylum",
  "Subphylum_1", "Superclass", "Class", "Subclass", "Infraclass",
  "Superorder", "Order", "Suborder_1", "Suborder_2", "Suborder_3",
  "Family", "Subfamily", "Genus", "Subgenus", "Species"
)

clean_mzg_taxonomy <- function(taxa) {
  taxa <- as.matrix(taxa)
  colnames(taxa) <- mzg_ranks[seq_len(ncol(taxa))]
  taxa
}

add_dna_sequences <- function(ps) {
  dna <- DNAStringSet(taxa_names(ps))
  names(dna) <- taxa_names(ps)
  merge_phyloseq(ps, dna)
}

build_marker_phyloseq <- function(datadir, metadata_file, output_file) {
  seqtab <- readRDS(file.path(datadir, "seqtab_nochim.rds"))
  taxa <- readRDS(file.path(datadir, "taxa.rds")) |> clean_mzg_taxonomy()
  metadata <- readRDS(file.path(datadir, metadata_file))

  rownames(seqtab) <- sub("_$", "", rownames(seqtab))

  ps <- phyloseq(
    otu_table(seqtab, taxa_are_rows = FALSE),
    sample_data(metadata),
    tax_table(taxa)
  ) |>
    add_dna_sequences()

  taxa_names(ps) <- paste0("ASV", seq_len(ntaxa(ps)))
  saveRDS(ps, output_file)
  ps
}

read_depth_plot <- function(ps, title, y_label = "Number of reads") {
  tibble(
    Sample = sample_names(ps),
    Reads = sample_sums(ps)
  ) |>
    ggplot(aes(Sample, Reads)) +
    geom_col() +
    scale_x_discrete(
      labels = \(x) substr(x, 1, 6),
      breaks = \(x) x[seq(1, length(x), by = 4)]
    ) +
    scale_y_continuous(labels = label_number(big.mark = ",")) +
    labs(title = title, x = "Sample", y = y_label) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

presence_absence <- function(ps) {
  otu_table(ps) <- otu_table((otu_table(ps) > 0) * 1, taxa_are_rows(ps))
  ps
}

present_taxa_at_rank <- function(ps, tax_rank) {
  tax <- as.data.frame(phyloseq::tax_table(ps), stringsAsFactors = FALSE)

  if (!tax_rank %in% colnames(tax)) {
    stop("Taxonomic rank not found: ", tax_rank)
  }

  present_ids <- phyloseq::taxa_names(ps)[phyloseq::taxa_sums(ps) > 0]

  taxa_values <- tax[present_ids, tax_rank, drop = TRUE]

  taxa_values <- taxa_values[
    !is.na(taxa_values) &
      taxa_values != "" &
      taxa_values != "NA" &
      taxa_values != "N/A" &
      taxa_values != "Unknown"
  ]

  unique(as.character(taxa_values))
}

phyloseq_to_pa_long <- function(ps, method, taxa_keep, tax_rank, lon_var = NULL) {
  otu <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) otu <- t(otu)
  otu <- (otu > 0) * 1

  tax <- as(tax_table(ps), "matrix") |>
    as.data.frame(check.names = FALSE) |>
    rownames_to_column("taxa_id")

  samples <- as(sample_data(ps), "data.frame") |>
    rownames_to_column("SampleID")

  if (is.null(lon_var)) {
    lon_var <- names(samples)[str_detect(str_to_lower(names(samples)), "^lon$|^longitude$|long")][1]
  }

  stopifnot(!is.na(lon_var), lon_var %in% names(samples), tax_rank %in% names(tax))

  as.data.frame(otu) |>
    rownames_to_column("taxa_id") |>
    pivot_longer(-taxa_id, names_to = "SampleID", values_to = "pa") |>
    left_join(tax, by = "taxa_id") |>
    left_join(samples, by = "SampleID") |>
    transmute(
      method = method,
      SampleID = SampleID,
      longitude = as.numeric(.data[[lon_var]]),
      taxon = as.character(.data[[tax_rank]]),
      pa = as.integer(pa > 0)
    ) |>
    filter(!is.na(taxon), taxon %in% taxa_keep)
}

binned_presence_plot <- function(
  ps_cpr_pa,
  ps_edna_pa,
  taxa_keep,
  tax_rank,
  palette,
  lon_min = -16.5,
  lon_max = -1,
  n_bins = 10
) {
  lon_breaks <- seq(lon_min, lon_max, length.out = n_bins + 1)

  plot_df <- bind_rows(
    phyloseq_to_pa_long(ps_cpr_pa, "CPR", taxa_keep, tax_rank),
    phyloseq_to_pa_long(ps_edna_pa, "eDNA", taxa_keep, tax_rank)
  ) |>
    filter(!is.na(longitude), between(longitude, lon_min, lon_max)) |>
    mutate(lon_bin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE)) |>
    group_by(method, lon_bin, taxon) |>
    summarise(pa = as.integer(any(pa == 1)), .groups = "drop") |>
    mutate(method = factor(method, levels = c("CPR", "eDNA")))

  taxon_levels <- plot_df |>
    group_by(taxon) |>
    summarise(prevalence = sum(pa), .groups = "drop") |>
    arrange(desc(prevalence), taxon) |>
    pull(taxon)

  bin_mids <- (lon_breaks[-1] + lon_breaks[-length(lon_breaks)]) / 2
  mid_labels <- sprintf("%.2f", bin_mids)
  names(mid_labels) <- levels(plot_df$lon_bin)

  plot_df |>
    mutate(
      taxon = factor(taxon, levels = taxon_levels),
      fill_taxon = if_else(pa == 1, as.character(taxon), NA_character_)
    ) |>
    ggplot(aes(lon_bin, taxon, fill = fill_taxon)) +
    geom_tile(width = 0.95, height = 0.9, colour = "grey80") +
    facet_wrap(~ method, ncol = 1) +
    scale_fill_manual(
      values = palette[taxon_levels],
      na.value = "white",
      breaks = taxon_levels,
      limits = taxon_levels,
      drop = FALSE,
      name = tax_rank
    ) +
    scale_x_discrete(labels = mid_labels) +
    scale_y_discrete(limits = rev(taxon_levels)) +
    labs(x = "Longitude section, bin midpoint", y = NULL) +
    theme(
      panel.grid = element_blank(),
      strip.text = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
}

calculate_taxon_prevalence <- function(ps, taxa_keep, tax_rank = "Genus", method = NULL) {
  otu <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) otu <- t(otu)

  tax <- as(tax_table(ps), "matrix") |>
    as.data.frame(check.names = FALSE) |>
    rownames_to_column("taxa_id")

  out <- as.data.frame(otu > 0) |>
    rownames_to_column("taxa_id") |>
    pivot_longer(-taxa_id, names_to = "SampleID", values_to = "present") |>
    left_join(tax, by = "taxa_id") |>
    mutate(taxon = as.character(.data[[tax_rank]])) |>
    filter(!is.na(taxon), taxon %in% taxa_keep) |>
    group_by(SampleID, taxon) |>
    summarise(present = any(present), .groups = "drop") |>
    complete(SampleID = sample_names(ps), taxon = taxa_keep, fill = list(present = FALSE)) |>
    group_by(taxon) |>
    summarise(
      n_present = sum(present),
      n_samples = n_distinct(SampleID),
      prevalence = 100 * n_present / n_samples,
      .groups = "drop"
    ) |>
    arrange(desc(prevalence))

  if (!is.null(method)) out <- mutate(out, method = method)
  out
}

community_matrix <- function(ps, taxrank = NULL) {
  if (!is.null(taxrank)) {
    ps <- tax_glom(ps, taxrank = taxrank, NArm = TRUE)
  }

  mat <- as(otu_table(ps), "matrix")
  if (taxa_are_rows(ps)) mat <- t(mat)

  mat <- (mat > 0) * 1
  mat <- mat[rowSums(mat) > 0, colSums(mat) > 0, drop = FALSE]
  mat
}

nmds_plot <- function(mat, metadata, title = NULL) {
  nmds <- metaMDS(
    mat,
    distance = "jaccard",
    binary = TRUE,
    k = 2,
    trymax = 200,
    autotransform = FALSE,
    trace = FALSE
  )

  scores(nmds, display = "sites") |>
    as.data.frame() |>
    rownames_to_column("SampleID") |>
    left_join(metadata, by = "SampleID") |>
    ggplot(aes(NMDS1, NMDS2, colour = Water_masses)) +
    geom_point(size = 3, alpha = 0.7) +
    stat_ellipse(aes(group = Water_masses), linewidth = 0.8, linetype = "dashed") +
    scale_colour_manual(values = c("Open Ocean" = "#D55E00", "Shelf" = "#0072B2")) +
    coord_equal(xlim = c(-1.3, 1.0), ylim = c(-1.3, 1.0)) +
    labs(title = title, colour = NULL)
}
```

# Read tracking

```{r read-tracking}
dada2_counts <- read_csv("data/DADA2_counts_table.csv")

colnames(dada2_counts) <- c(
  "sample",
  "input", "filtered", "nonchim", "reads_retained (%)",
  "input", "filtered", "nonchim", "reads_retained (%)"
)

kable(dada2_counts, format = "html", caption = "DADA2 read counts for 18S and COI") |>
  add_header_above(c(" " = 1, "18S" = 4, "COI" = 4)) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed")
  ) |>
  scroll_box(height = "400px", width = "100%")
```

::: {.callout-note title="Negative controls"}
The no-template control and extraction blank in the 18S dataset did not pass the initial DADA2 filtering step. These samples therefore had zero retained reads after quality filtering and were excluded from downstream denoising, merging, and taxonomic assignment.
:::

# Build phyloseq objects

## 18S

```{r build-18s}
ps_18s <- build_marker_phyloseq(
  datadir = "data/18S",
  metadata_file = "metadata_18S.rds",
  output_file = "data/ps_18s.rds"
)

ps_18s
sample_variables(ps_18s)
read_depth_plot(ps_18s, "18S sequencing depth per sample", "Number of non-chimeric reads")
```

## COI

```{r build-coi}
ps_coi <- build_marker_phyloseq(
  datadir = "data/COI",
  metadata_file = "metadata_COI.rds",
  output_file = "data/ps_coi.rds"
)

ps_coi
sample_variables(ps_coi)
read_depth_plot(ps_coi, "COI sequencing depth per sample", "Number of non-chimeric reads")
```

## CPR

```{r build-cpr}
counts <- read_csv("data/count_table_CPR.csv")
taxo <- read_csv("data/taxonomy_table_CPR.csv")
meta <- read_csv("data/metadata_CPR.csv")

taxtab <- as.matrix(taxo)
rownames(taxtab) <- taxo[[8]]

ps_cpr <- phyloseq(
  otu_table(counts, taxa_are_rows = FALSE),
  sample_data(meta),
  tax_table(taxtab)
)

saveRDS(ps_cpr, "data/ps_cpr.rds")

ps_cpr
sample_variables(ps_cpr)
rank_names(ps_cpr)
read_depth_plot(ps_cpr, "CPR counts per sample", "Number of CPR counts")

meta_cpr <- as(sample_data(ps_cpr), "data.frame") |>
  rownames_to_column("SampleID")

```
# Map

```{r map}

gps <- read_xlsx("data/CPR_coordinates.xlsx")

cols <- c(
  "1" = "#1B9E77",
  "2" = "#D95F02",
  "3" = "#7570B3",
  "4" = "#E7298A"
)

library(here)

options(
  ggOceanMaps.datapath = here(".ggOceanMaps_data")
)

dir.create(
  getOption("ggOceanMaps.datapath"),
  recursive = TRUE,
  showWarnings = FALSE
)

p_map <- basemap(
  limits = c(-18, 0, 46, 52.5),
  bathy.style = "rcb",
  land.col = "grey80",
  land.border.col = "grey35",
  bathy.border.col = NA
) +
  
  # tone down bathymetry
  scale_fill_gradientn(
    colours = c("#f7fbff", "#deebf7", "#c6dbef", "#9ecae1", "#6baed6"),
    na.value = "white",
    name = "Depth (m)",
    breaks = c(0, 1000, 2000, 3000, 4000, 5000),
    labels = c("0", "1000", "2000", "3000", "4000", "5000"),
    guide = guide_colorbar(
      title.position = "top",
      title.hjust = 0,
      barheight = unit(3.5, "cm"),
      barwidth = unit(0.35, "cm"),
      frame.colour = "grey50",
      ticks.colour = "grey40"
    )
  ) +
  
  ggnewscale::new_scale_fill() +
  
  geom_spatial_point(
    data = meta_cpr,
    aes(
      x = Longitude,
      y = Latitude
    ),
    colour = "black",
    size = 3,
    shape= 4,
    stroke = 1.2
  ) +
  
  geom_spatial_point(
    data = gps,
    aes(
      x = Lon,
      y = Lat,
      fill = factor(Transect, levels = c(1, 2, 3, 4))
    ),
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 2.1,
    alpha = 0.95
  ) +
  
  scale_fill_manual(
    values = cols,
    name = "Sampling\ntransect",
    labels = c("1", "2", "3", "4")
  ) +
  
  guides(
    fill = guide_legend(
      override.aes = list(size = 3, stroke = 0.3),
      title.position = "top",
      title.hjust = 0
    )
  ) +
  
  annotation_scale(
    location = "br",
    width_hint = 0.25,
    text_cex = 0.7,
    line_width = 0.4,
    pad_x = unit(0.25, "in"),
    pad_y = unit(0.18, "in")
  ) +
  
  annotation_north_arrow(
    location = "br",
    which_north = "true",
    pad_x = unit(0.35, "in"),
    pad_y = unit(0.55, "in"),
    style = north_arrow_fancy_orienteering(
      text_size = 7,
      line_width = 0.35
    )
  ) +
  
  theme_minimal(base_size = 10) +
  theme(
    axis.title = element_blank(),
    axis.text = element_text(size = 9, colour = "grey20"),
    
    panel.grid.major = element_line(linewidth = 0.25, colour = "grey75"),
    panel.grid.minor = element_blank(),
    
    legend.position = c(0.08, 0.72),
    legend.box = "vertical",
    legend.background = element_rect(
      fill = alpha("white", 0.75),
      colour = NA
    ),
    legend.key = element_rect(
      fill = NA,
      colour = NA
    ),
    legend.key.size = unit(0.45, "cm"),
    
    legend.title = element_text(size = 9, face = "bold"),
    legend.text = element_text(size = 8.5),
    
    panel.border = element_rect(
      fill = NA,
      colour = "grey25",
      linewidth = 0.4
    ),
    
    plot.background = element_rect(fill = "white", colour = NA),
    plot.margin = margin(4, 4, 4, 4)
  )

### map with CPR samples as a cross

p_map2 <- p_map +
  geom_spatial_point(
    data = meta_cpr,
    aes(
      x = Longitude,
      y = Latitude
    ),
    colour = "black",
    size = 3,
    shape= 4
  )

p_map2
```



# Merge 18S and COI eDNA datasets

```{r merge-edna}
ps_18s <- readRDS("data/ps_18s.rds")
ps_coi <- readRDS("data/ps_coi.rds")
ps_cpr <- readRDS("data/ps_cpr.rds")

# Ensure ASV identifiers are unique across markers.
taxa_names(ps_18s) <- paste0("18S__", taxa_names(ps_18s))
taxa_names(ps_coi) <- paste0("COI__", taxa_names(ps_coi))

add_marker <- function(ps, marker) {
  tax <- as(tax_table(ps), "matrix")
  tax <- cbind(tax, marker = marker)
  tax_table(ps) <- tax_table(tax)
  ps
}

pad_tax_table <- function(ps, all_ranks) {
  tax <- as(tax_table(ps), "matrix")
  missing_ranks <- setdiff(all_ranks, colnames(tax))

  if (length(missing_ranks) > 0) {
    tax <- cbind(
      tax,
      matrix(
        NA,
        nrow = nrow(tax),
        ncol = length(missing_ranks),
        dimnames = list(rownames(tax), missing_ranks)
      )
    )
  }

  tax_table(ps) <- tax_table(tax[, all_ranks, drop = FALSE])
  ps
}

ps_18s <- add_marker(ps_18s, "18S")
ps_coi <- add_marker(ps_coi, "COI")
all_ranks <- union(colnames(tax_table(ps_18s)), colnames(tax_table(ps_coi)))

ps_edna <- merge_phyloseq(
  pad_tax_table(ps_18s, all_ranks),
  pad_tax_table(ps_coi, all_ranks)
)

ps_edna
table(tax_table(ps_edna)[, "marker"])
```

# Harmonising taxonomy

Taxonomic nomenclature was reviewed prior to comparison of the CPR and eDNA datasets. Historical synonyms were harmonised where appropriate.

```{r taxonomic-harmonisation}

# Extract taxonomy
tax_cpr <- as.data.frame(tax_table(ps_cpr), stringsAsFactors = FALSE)

# CPR
tax_cpr$Genus <- dplyr::recode(
  tax_cpr$Genus,
  "Ceratium" = "Tripos",
  "Emiliania" = "Gephyrocapsa",
  "Corycaeus" = "Ditrichocorycaeus"
)

tax_cpr[tax_cpr$Genus == "Rhizomonas / Solenicola", ] <- within(
  tax_cpr[tax_cpr$Genus == "Rhizomonas / Solenicola", ],
  {
    Kingdom <- "Chromista"
    Phylum  <- "Bigyra"
    Class   <- "Nanomonadea"
    Order   <- "Uniciliatida"
    Family  <- "Solenicolidae"
    Genus   <- "Solenicola"
    Species <- "setigera"
    Organism <- "Solenicola setigera"
  }
)

# eDNA
tax_edna <- as.data.frame(tax_table(ps_edna), stringsAsFactors = FALSE)

tax_edna$Genus <- dplyr::recode(
  tax_edna$Genus,
  "Pseudonitzschia" = "Pseudo-nitzschia"
)

# Fix Species name issue

tax_cpr <- tax_cpr |>
  mutate(
    Species_clean = Species |>
      str_remove_all("\\s*\\([^)]*\\)") |>   # remove anything in parentheses
      str_remove_all("\\b[Nn]auplii\\b") |>  # remove "nauplii"
      str_remove_all("\\bcomplex\\b") |>     # remove "complex"
      str_squish(),                          # remove extra spaces

    Species_full = case_when(
      Species_clean %in% c("N/A", "", NA) ~ "N/A",
      TRUE ~ paste(Genus, Species_clean, sep = "_")
    ),

    Species_full_clean = if_else(
      str_detect(Species_full, "^[A-Z][A-Za-z-]+_[a-z-]+$"),
      Species_full,
      NA_character_
    )
  )

length(unique(na.omit(tax_cpr$Species_full_clean)))
cpr_sp <- unique(na.omit(tax_cpr$Species_full_clean))
edna_sp <- unique(na.omit(tax_edna$Species))
intersect(cpr_sp, edna_sp)

tax_cpr$Species <- tax_cpr$Species_full_clean
tax_table(ps_edna) <- tax_table(as.matrix(tax_edna))
tax_table(ps_cpr) <- tax_table(as.matrix(tax_cpr))

```

Genus names were harmonised between datasets to reflect current accepted taxonomy. This included correcting database-specific spellings (e.g. Pseudonitzschia to Pseudo-nitzschia) and incorporating recent taxonomic revisions (Ceratium → Tripos, Emiliania → Gephyrocapsa, Corycaeus → Ditrichocorycaeus). This ensures that future readers understand these are deliberate nomenclatural standardisations rather than modifications to the biological data.

# Assessment of apparent CPR-only genera

Proboscia and Dictyocysta were highly prevalent in the CPR dataset but absent from the eDNA genus-level assignments, despite both genera being represented in the 18S reference database. BLAST analysis of unassigned Rhizosoleniaceae and Tintinnidae ASVs returned closest matches to Rhizosolenia and Eutintinnus, respectively, providing no evidence that these sequences belonged to Proboscia or Dictyocysta. These mismatches are therefore unlikely to result solely from incomplete reference database coverage and may instead reflect limitations in taxonomic assignment, marker or primer bias, or genuine differences in taxon detection between CPR microscopy and eDNA metabarcoding.

Examination of the COI reference database indicated that several CPR-only copepod genera (Lepas, Paraeuchaeta, Aetideus, Scina and Oikopleura) were represented by reference sequences, suggesting that their absence from the eDNA dataset is unlikely to result from incomplete database coverage. In contrast, Ischnocalanus and Parapontella were absent from the reference database, while Alteutha was represented by a single sequence, indicating that limited reference coverage may contribute to their apparent absence. Finally, Pontellidae and Euchaetidae are family-level CPR categories, and Para-Pseudocalanus represents a CPR-specific operational taxonomic group rather than a standard genus.

Several CPR-only copepod genera (e.g. Paraeuchaeta, Aetideus, Scina and Ischnocalanus) are relatively large, low-abundance taxa. Their absence from the eDNA dataset may therefore reflect differences in sampling volume rather than limitations of metabarcoding per se. Each CPR sample integrates approximately 3 m³ of seawater, whereas eDNA analyses were based on ~2 L water samples, resulting in a >1,000-fold difference in sampled volume. Such differences are expected to disproportionately reduce the probability of detecting large, sparsely distributed metazoans in eDNA samples.

# Taxonomic overlap between eDNA and CPR

```{r taxonomic-overlap}
ps_edna_pa <- presence_absence(ps_edna)
ps_cpr_pa <- presence_absence(ps_cpr)

ranks_to_test <- intersect(rank_names(ps_edna), rank_names(ps_cpr))

taxa_overlap_summary <- map_dfr(ranks_to_test, \(tax_rank) {
  edna_taxa <- present_taxa_at_rank(ps_edna_pa, tax_rank)
  cpr_taxa  <- present_taxa_at_rank(ps_cpr_pa, tax_rank)

  n_shared <- length(intersect(edna_taxa, cpr_taxa))

  tibble(
    rank = tax_rank,
    n_edna = length(edna_taxa),
    n_cpr = length(cpr_taxa),
    n_shared = n_shared,
    perc_shared_edna = round(100 * n_shared / n_edna, 1),
    perc_shared_cpr = round(100 * n_shared / n_cpr, 1)
  )
})

kable(
  taxa_overlap_summary,
  caption = "Taxonomic overlap between eDNA and CPR datasets, based on presence/absence.",
  digits = 1
) |>
  kable_styling(full_width = FALSE)
```

# Longitudinal prevalence plots

## Shared zooplankton genera

```{r zooplankton-longitude-plot, fig.width=8, fig.height=6}
zooplankton_genera <- c(
  "Paracalanus", "Clausocalanus", "Nannocalanus", "Centropages", "Acartia",
  "Calanus", "Temora", "Metridia", "Oithona", "Oncaea"
)

zooplankton_cols <- setNames(
  c("#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A", "#984EA3",
    "#A6D854", "#FFFF33", "#FF7F00", "#F781BF", "#999999"),
  zooplankton_genera
)

binned_presence_plot(
  ps_cpr_pa = ps_cpr_pa,
  ps_edna_pa = ps_edna_pa,
  taxa_keep = zooplankton_genera,
  tax_rank = "Genus",
  palette = zooplankton_cols
)
```

## Shared phytoplankton families

```{r phytoplankton-longitude-plot, fig.width=8, fig.height=6}
phytoplankton_families <- c(
  "Rhizosoleniaceae", "Chaetoceraceae", "Ceratiaceae", "Coccolithaceae",
  "Dictyochaceae", "Leptocylindraceae", "Noelaerhabdaceae",
  "Rhabdosphaeraceae", "Prorocentraceae", "Thalassiosiraceae"
)

phytoplankton_cols <- setNames(
  c("#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#FFD92F",
    "#E5C494", "#B3B3B3", "#A65628", "#FFFFB3", "#008B8B"),
  phytoplankton_families
)

binned_presence_plot(
  ps_cpr_pa = ps_cpr_pa,
  ps_edna_pa = ps_edna_pa,
  taxa_keep = phytoplankton_families,
  tax_rank = "Family",
  palette = phytoplankton_cols
)
```

### Bin membership table

```{r bin membership}
lon_min <- -16.5
lon_max <- -1
n_bins  <- 10

lon_breaks <- seq(lon_min, lon_max, length.out = n_bins + 1)

cpr_meta <- data.frame(sample_data(ps_cpr_pa)) |>
  rownames_to_column("sample_id") |>
  mutate(
    Method = "CPR",
    longitude = Longitude,
    latitude = Latitude
  )

edna_meta <- data.frame(sample_data(ps_edna_pa)) |>
  rownames_to_column("sample_id") |>
  mutate(
    Method = "eDNA",
    longitude = lon,
    latitude = lat
  )

sample_bins <- bind_rows(cpr_meta, edna_meta) |>
  filter(!is.na(longitude), between(longitude, lon_min, lon_max)) |>
  mutate(
    Bin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE),
    Bin_midpoint = (lon_breaks[as.integer(Bin)] + lon_breaks[as.integer(Bin) + 1]) / 2
  ) |>
  arrange(Bin, Method, longitude)

# Samples binned together
bin_table <- sample_bins |>
  group_by(Bin, Bin_midpoint, Method) |>
  summarise(
    n_samples = n(),
    samples = paste(sample_id, collapse = ", "),
    numbers = paste(Number, collapse = ", "),
    lon_range = paste(round(min(longitude), 3), "to", round(max(longitude), 3)),
    .groups = "drop"
  )

kable(bin_table, format = "html", caption = "Bin membership table")

```

# Genus prevalence comparison

```{r genus-prevalence-scatter, fig.width=12, fig.height=8}
shared_genera <- intersect(
  present_taxa_at_rank(ps_edna_pa, "Genus"),
  present_taxa_at_rank(ps_cpr_pa, "Genus")
)

assign_broad_group <- function(ps, genera_keep) {
  as(tax_table(ps), "matrix") |>
    as.data.frame(check.names = FALSE) |>
    mutate(across(everything(), as.character)) |>
    filter(!is.na(Genus), Genus %in% genera_keep) |>
    distinct(Genus, Phylum, Class) |>
    mutate(
      broad_group = case_when(
        Phylum %in% c(
          "Arthropoda", "Cnidaria", "Ctenophora", "Mollusca", "Annelida",
          "Chaetognatha", "Chordata", "Rotifera", "Ciliophora"
        ) ~ "Zooplankton",
        Phylum %in% c(
          "Bacillariophyta", "Dinoflagellata", "Haptophyta", "Ochrophyta",
          "Chlorophyta", "Cryptophyta", "Cyanobacteria"
        ) |
          Class %in% c(
            "Mediophyceae", "Bacillariophyceae", "Coscinodiscophyceae",
            "Dinophyceae", "Prymnesiophyceae", "Dictyochophyceae",
            "Chlorophyceae", "Cryptophyceae"
          ) ~ "Phytoplankton",
        TRUE ~ "Other/ambiguous"
      )
    ) |>
    group_by(Genus) |>
    summarise(
      broad_group = case_when(
        any(broad_group == "Zooplankton") ~ "Zooplankton",
        any(broad_group == "Phytoplankton") ~ "Phytoplankton",
        TRUE ~ "Other/ambiguous"
      ),
      .groups = "drop"
    )
}

genus_prev <- bind_rows(
  calculate_taxon_prevalence(ps_cpr_pa, shared_genera, "Genus", "CPR"),
  calculate_taxon_prevalence(ps_edna_pa, shared_genera, "Genus", "eDNA")
) |>
  dplyr::rename(Genus = taxon) |>
  dplyr::select(Genus, method, prevalence) |>
  pivot_wider(names_from = method, values_from = prevalence)

genus_groups <- bind_rows(
  assign_broad_group(ps_cpr_pa, shared_genera),
  assign_broad_group(ps_edna_pa, shared_genera)
) |>
  group_by(Genus) |>
  summarise(
    broad_group = case_when(
      any(broad_group == "Zooplankton") ~ "Zooplankton",
      any(broad_group == "Phytoplankton") ~ "Phytoplankton",
      TRUE ~ "Other/ambiguous"
    ),
    .groups = "drop"
  )

plot_df <- genus_prev |>
  left_join(genus_groups, by = "Genus") |>
  arrange(desc(eDNA - CPR)) |>
  mutate(taxon_id = row_number())

cor_test <- cor.test(plot_df$CPR, plot_df$eDNA, method = "spearman")

p_scatter <- ggplot(plot_df, aes(CPR, eDNA)) +
  annotate("polygon", x = c(0, 100, 0), y = c(0, 100, 100), fill = "#DCEAF7", alpha = 0.4) +
  annotate("polygon", x = c(0, 100, 100), y = c(0, 0, 100), fill = "#EEEEEE", alpha = 0.6) +
  annotate("text", x = 55, y = 100, label = "Higher prevalence\nin eDNA", colour = "#3B6FB6", fontface = "italic", hjust = 0, vjust = 1, size = 4) +
  annotate("text", x = 73, y = 60, label = "Higher prevalence\nin CPR", colour = "grey40", fontface = "italic", hjust = 0, size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55", linewidth = 0.5) +
  geom_point(aes(fill = broad_group), shape = 21, size = 6.5, colour = "black", stroke = 0.45, alpha = 0.95) +
  geom_text(aes(label = taxon_id), colour = "white", fontface = "bold", size = 3.2) +
  scale_fill_manual(values = c("Zooplankton" = "#E69F00", "Phytoplankton" = "#0072B2", "Other/ambiguous" = "grey60")) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25), expand = expansion(mult = c(0.02, 0.03))) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25), expand = expansion(mult = c(0.02, 0.03))) +
  coord_equal(clip = "off") +
  labs(
    x = "Prevalence in CPR samples (%)",
    y = "Prevalence in eDNA samples (%)",
    title = "Comparison of genus prevalence between CPR and eDNA",
    subtitle = paste0(
      "n = ", nrow(plot_df),
      "; Spearman rho = ", round(unname(cor_test$estimate), 2),
      "; p = ", signif(cor_test$p.value, 2)
    )
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.4)
  )

taxa_table <- plot_df |>
  arrange(taxon_id) |>
  select(taxon_id, Genus)

table_plot <- arrangeGrob(
  tableGrob(taxa_table[1:ceiling(nrow(taxa_table) / 2), ], rows = NULL),
  tableGrob(taxa_table[(ceiling(nrow(taxa_table) / 2) + 1):nrow(taxa_table), ], rows = NULL),
  ncol = 2
)

p_scatter + wrap_elements(table_plot) + plot_layout(widths = c(2.6, 1.4))
```

# Community ordination

## CPR, presence/absence Jaccard NMDS

```{r cpr-nmds}
sample_data(ps_cpr)$Water_masses <- if_else(
  sample_data(ps_cpr)$Longitude < -10,
  "Open Ocean",
  "Shelf"
) |>
  factor(levels = c("Open Ocean", "Shelf"))

mat_cpr_pa <- community_matrix(ps_cpr)
meta_cpr <- as(sample_data(ps_cpr), "data.frame") |>
  rownames_to_column("SampleID") |>
  filter(SampleID %in% rownames(mat_cpr_pa))

nmds_plot(mat_cpr_pa, meta_cpr, title = "CPR microscopy")
```

```{r cpr-permanova}
adonis2(
  mat_cpr_pa ~ Water_masses,
  data = meta_cpr,
  method = "jaccard",
  binary = TRUE,
  permutations = 999
)
```

```{r cpr-betadisper}
dist_cpr <- vegdist(mat_cpr_pa, method = "jaccard", binary = TRUE)
disp_cpr <- betadisper(dist_cpr, meta_cpr$Water_masses)

anova(disp_cpr)
permutest(disp_cpr)
```

## eDNA, genus-level presence/absence Jaccard NMDS

```{r edna-genus-nmds}
sample_data(ps_edna)$Water_masses <- if_else(
  sample_data(ps_edna)$lon < -10,
  "Open Ocean",
  "Shelf"
) |>
  factor(levels = c("Open Ocean", "Shelf"))

mat_edna_genus_pa <- community_matrix(ps_edna, taxrank = "Genus")
meta_edna <- as(sample_data(ps_edna), "data.frame") |>
  rownames_to_column("SampleID") |>
  filter(SampleID %in% rownames(mat_edna_genus_pa))

nmds_plot(mat_edna_genus_pa, meta_edna, title = "eDNA metabarcoding, genus level")
```

```{r edna-genus-permanova}
adonis2(
  mat_edna_genus_pa ~ Water_masses,
  data = meta_edna,
  method = "jaccard",
  binary = TRUE,
  permutations = 999
)
```

```{r edna-genus-betadisper}
dist_edna_genus <- vegdist(mat_edna_genus_pa, method = "jaccard", binary = TRUE)
disp_edna_genus <- betadisper(dist_edna_genus, meta_edna$Water_masses)

anova(disp_edna_genus)
permutest(disp_edna_genus)
```

# Genera unique to CPR or eDNA

```{r unique-genera}
edna_genera <- present_taxa_at_rank(ps_edna, "Genus")
cpr_genera <- present_taxa_at_rank(ps_cpr, "Genus")

shared_genera <- intersect(edna_genera, cpr_genera)
cpr_only_genera <- setdiff(cpr_genera, edna_genera)
edna_only_genera <- setdiff(edna_genera, cpr_genera)

tibble(
  category = c("Shared", "CPR only", "eDNA only"),
  n_genera = c(length(shared_genera), length(cpr_only_genera), length(edna_only_genera))
)
```

## Most prevalent CPR-only genera

```{r cpr-only-prevalence}
cpr_only_prev <- calculate_taxon_prevalence(ps_cpr, cpr_only_genera, "Genus") |>
  dplyr::rename(Genus = taxon) |>
  mutate(category = "CPR only")

cpr_only_prev |>
  slice_max(prevalence, n = 20)
```

## Most prevalent eDNA-only genera

```{r edna-only-prevalence}
edna_only_prev <- calculate_taxon_prevalence(ps_edna, edna_only_genera, "Genus") |>
  dplyr::rename(Genus = taxon) |>
  mutate(category = "eDNA only")

edna_only_prev |>
  slice_max(prevalence, n = 20)
```

## Taxonomic identity of eDNA-only genera

```{r edna-only-taxonomy}
edna_only_taxonomy <- as.data.frame(tax_table(ps_edna)) |>
  filter(Genus %in% edna_only_genera) |>
  select(Phylum, Class, Order, Family, Genus) |>
  distinct()

edna_only_summary <- edna_only_prev |>
  left_join(edna_only_taxonomy, by = "Genus") |>
  arrange(desc(prevalence))

head(edna_only_summary, 30)
```

```{r edna-only-classes, fig.width=8, fig.height=5}
edna_only_summary |>
  group_by(Class) |>
  summarise(
    n_genera = n_distinct(Genus),
    mean_prevalence = mean(prevalence),
    .groups = "drop"
  ) |>
  filter(mean_prevalence > 10, n_genera >= 5) |>
  arrange(desc(n_genera)) |>
  slice_head(n = 15) |>
  mutate(Class = forcats::fct_reorder(Class, n_genera)) |>
  ggplot(aes(n_genera, Class, fill = mean_prevalence)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = n_genera), hjust = -0.25, size = 3.5) +
  scale_fill_gradient(low = "grey85", high = "black", name = "Mean\nprevalence (%)") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "Number of genera detected only by eDNA",
    y = NULL,
    title = "Taxonomic classes containing genera uniquely detected by eDNA"
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```
