---
title: "Analysis Code for the RoCSI-CPR study"
author: "Nathan Hubot, Susan Evans, Julie Robidart"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
editor: source
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

# 🧬 Overview of Sequencing & Experimental Design

This report summarises the **two metabarcoding workflows** applied to the RoCSI-CPR eDNA samples:

-   **18S rRNA V9 region** (targeting phytoplankton)
-   **COI mitochondrial region** (targeting zooplankton)

Both datasets were sequenced at the\
**Centre for Genomic Research (CGR), University of Liverpool.**

## 📊 Sequencing Summary

| Feature | **18S rRNA V9** | **COI (metazoans)** |
|-----------------|-------------------------|-------------------------------|
| **SSP ID** | [SSP202877](https://cgr.liv.ac.uk/illum/SSP202877_b388934ddf413642/) | [SSP200993](https://cgr.liv.ac.uk/illum/SSP200993_bbba6061877a5e12) |
| **Purchase order** | 203631529 | P10817-5 |
| **Sequencing platform** | Illumina MiSeq v2 (2×150 bp) | Illumina MiSeq v2 (2×250 bp) |
| **Target region** | 18S V9 (1389F–1510R) | COI Leray fragment (m1COIintF–jgHCO2198) |
| **Expected amplicon size** | \~121 bp | \~313 bp |
| **Raw read depth (mean)** | \~150k | \~200k |
| **Pre-trimming QC** | Cutadapt v4.5 | Cutadapt v1.2.1 + Sickle v1.200 |
| **Downstream processing** | Cutadapt → DADA2 → MZG 18S | Cutadapt → DADA2 → MZG COI |

------------------------------------------------------------------------

# 🧬 Primer Sets (with CGR Overhangs)

## 18S V9 Primers

-   **Forward:**\
    5′ [ACACTCTTTCCCTACACGACGCTCTTCCGATCTNNNNN]{style="color:#2E86C1;font-weight:bold;"}\
    [TTGTACACACCGCCC]{style="color:#C0392B;font-weight:bold;"} 3′\
    *(CGR overhang + spacer in blue; 18S primer in red)*

-   **Reverse:**\
    5′ [GTGACTGGAGTTCAGACGTGTGCTCTTCCGATCT]{style="color:#2E86C1;font-weight:bold;"}\
    [CCTTCYGCAGGTTCACCTAC]{style="color:#C0392B;font-weight:bold;"} 3′

------------------------------------------------------------------------

## COI Primers

-   **Forward:**\
    5′ [ACACTCTTTCCCTACACGACGCTCTTCCGATCTNNNNN]{style="color:#2E86C1;font-weight:bold;"}\
    [GGWACWGGWTGAACWGTWTAYCCYCC]{style="color:#C0392B;font-weight:bold;"} 3′\
    *(m1COIintF)*

-   **Reverse:**\
    5′ [GTGACTGGAGTTCAGACGTGTGCTCTTCCGATCT]{style="color:#2E86C1;font-weight:bold;"}\
    [TAIACYTCIGGRTGICCRAARAAYCA]{style="color:#C0392B;font-weight:bold;"} 3′\
    *(jgHCO2198)*

------------------------------------------------------------------------

# 🔧 Bioinformatic Workflow Overview

Both markers follow the same general pipeline:

1.  **Primer removal** (Cutadapt)\
2.  **Quality filtering** (DADA2)\
3.  **Error learning and denoising** (DADA2)\
4.  **Read merging** (DADA2)\
5.  **Chimera removal** (DADA2)\
6.  **Taxonomic assignment: ** (DADA2)
    -   18S → *MZG 18S “All Microbes + Protists”, Mode-A* reference database\
    -   COI → *MZG COI “All Invertebrates”, Mode-A* reference database

For each step, differences between the two pipelines are shown.

------------------------------------------------------------------------

# 1️⃣ Primer Removal with Cutadapt

### 🔹 Summary of Cutadapt parameters used

| Parameter | **18S** | **COI** |
|--------------------------|-----------------------|-----------------------|
| Forward primer (`-g`) | TTGTACACACCGCCC | GGWACWGGWTGAACWGTWTAYCCYCC |
| Reverse primer (`-G`) | CCTTCYGCAGGTTCACCTAC | TANACYTCNGGRTGNCCRAARAAYCA |
| `--match-read-wildcards` | ✔️ | ✔️ |
| Minimum overlap | **10** | **20** |
| Max error rate (`-e`) | **0.15** | **0.20** |
| `--discard-untrimmed` | ✔️ | ✔️ |
| `--minimum-length` | 80 bp | 200 bp |
| Pre-processing by CGR | Adapters trimmed via Cutadapt 4.5 | Adapters trimmed (Cutadapt v1.2.1) + Sickle quality filtering |

------------------------------------------------------------------------

### 🔹 Unified Cutadapt description

Both datasets used a looping bash command of the form:

``` bash

cutadapt\
  -g <forward_primer> \
  -G <reverse_primer> \
  --match-read-wildcards \
  --overlap <OV> \
  -e <ERROR> \
  --pair-filter=both \
  --discard-untrimmed \
  --cores=0 \
  -o $out1 -p $out2 \
  $f $r

```
Where <OV> and <ERROR> differ between markers (see table above).

# 2️⃣ DADA2 Processing

### 🔹 Summary of Cutadapt parameters used

| Parameters                | **18S**    | **COI**                             |
| ------------------------- | ---------- | ----------------------------------- |
| `truncLen`                | c(130,120) | c(210,210)                          |
| `maxEE`                   | c(2,3)     | c(2,3)                              |
| `minLen`                  | 80         | 200                                 |
| `minOverlap` (mergePairs) | 30         | 90                                  |
| Ref database              | MZG 18S    | MZG COI                             |

::: {.callout-note title="MZG Reference databases"}
| Marker  | Database                               | Notes                                        |
| ------- | -------------------------------------- | -------------------------------------------- |
| **18S** | `MZGdada2-18s__T2000000__o00__A.fastq` | phytoplankton --> "All Microbes + Protists", mode-A    |
| **COI** | `MZGdada2-coi__T4000000__o00__A.fastq` | zooplankton --> “All invertebrates”, mode-A          |
source: https://metazoogene.org/mzgdb/atlas/html-src/data__T4000000__o00.html
:::

### 🔹 Unified DADA2 description

Both datasets were processed with the standard DADA2 workflow:

``` r

# Filtering and trimming (R1/R2 after cutadapt)
filtered_out <- filterAndTrim(
  fwd  = forward_reads,
  filt = filtered_forward_reads,
  rev  = reverse_reads,
  filt.rev = filtered_reverse_reads,
  truncLen  = <MARKER_SPECIFIC>,   # see table above
  maxEE     = c(2, 3),             # expected errors (stricter for R1)
  maxN      = 0,                   # discard reads with Ns
  rm.phix   = TRUE,                # remove PhiX reads
  minLen    = <MINLEN>,            # marker-specific minimum length
  multithread = TRUE
)

# Error learning
errF <- learnErrors(filtered_forward_reads, multithread=TRUE)
errR <- learnErrors(filtered_reverse_reads, multithread=TRUE)

# Dereplication
derepF <- derepFastq(filtered_forward_reads)
derepR <- derepFastq(filtered_reverse_reads)

# ASV inference
dadaF <- dada(derepF, err=errF, pool="pseudo")
dadaR <- dada(derepR, err=errR, pool="pseudo")

# Merging
merged <- mergePairs(dadaF, derepF, dadaR, derepR,
                     minOverlap = <MARKER_SPECIFIC>,
                     trimOverhang = TRUE)

# ASV table
seqtab <- makeSequenceTable(merged)

# Chimera removal
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus")

# Taxonomic Assignment
taxa <- assignTaxonomy(
  seqtab.nochim,
  refFasta = <REF_FASTA>,
  multithread = TRUE
)

```

::: {.callout-note title="Note"}
For 18S, shorter reads (150 bp) and a very short amplicon (~121 bp) justify relatively short truncLen (130, 120) and minLen = 80.
For COI, the longer amplicon fragment (~313 bp) and 2×250 bp reads allow more aggressive truncation (210, 210) with large overlap (90) and a higher minLen = 200 to remove spurious short fragments.
:::

# 🔍 Read Tracking Summary

The following table summarizes read counts at each step of the DADA2 pipeline:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(knitr)
library(kableExtra)

# Load the data
df <- read_csv("data/DADA2_counts_table.csv")

# Rename columns to avoid ...2 suffix
colnames(df) <- c("sample",
                  "input", "filtered", "nonchim", "reads_retained (%)",
                  "input", "filtered", "nonchim", "reads_retained (%)")

# Create a styled table with group headers and scrolable
kable(df, format = "html", caption = "DADA2 Counts for 18S and COI") %>%
  add_header_above(c(" " = 1, "18S" = 4, "COI" = 4)) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed")
  ) %>%
  scroll_box(height = "400px", width = "100%")
```

::: {.callout-note title="Negative Controls"}
The **No-Template Control (NTC)** and **Extraction Blank** samples in the 18S dataset did not pass the initial **DADA2 filtering step**, resulting in zero retained reads after quality filtering.\
Consequently, these controls were **excluded from downstream analysis** (denoising, merging, and taxonomy assignment).

Their exclusion is consistent with expectations for negative controls, indicating the absence of detectable contamination above sequencing background levels.
:::

::: {.callout-note}
### 💬 Discussion — reads retained

The proportion of reads that were kept following the DADA2 pipeline is good: **84%** and **91%** for **18S** and **COI**, respectively. This leaves an average of **136K** and **142K** reads for **18S** and **COI**, respectively.
Let's note that for 18S, samples 12 and 13 lost more then the rest: **53%** and **60%**, respectively.
:::

# 18S Dataset

## 📦 Build the phyloseq object (18S)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)

# ---- paths ----
datadir <- "data/18S"

# ---- load DADA2 outputs ----
seqtab.nochim <- readRDS(file.path(datadir, "seqtab_nochim.rds"))
taxa          <- readRDS(file.path(datadir, "taxa.rds"))

# Make sure taxonomy is a proper matrix and has consistent column names
taxa <- as.matrix(taxa)
target_cols <- c(
  "Kingdom","Subkingdom_1","Subkingdom_2","Phylum","Subphylum","Subphylum_1","Superclass",
  "Class","Subclass","Infraclass","Superorder","Order","Suborder_1","Suborder_2","Suborder_3",
  "Family","Subfamily","Genus","Subgenus","Species"
)
colnames(taxa) <- target_cols[seq_len(ncol(taxa))]

# Load metadata
metadat <- readRDS(file.path(datadir, "metadata_18S.rds")) 

# Create a Phyloseq object

rownames(seqtab.nochim) <- sub("_$", "", rownames(seqtab.nochim))

ps_18s <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = F), 
               sample_data(metadat), 
               tax_table(taxa))
saveRDS(ps_18s, "data/ps_18s.rds")
  
cat("🔹 A summary of the phyloseq object \n")
ps_18s

# storing the ASV sequences and assigning them with a name and then renaming the taxa object
dna <- Biostrings::DNAStringSet(taxa_names(ps_18s))
names(dna) <- taxa_names(ps_18s)
ps_18s <- merge_phyloseq(ps_18s, dna)
taxa_names(ps_18s) <- paste0("ASV", seq(ntaxa(ps_18s)))

```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Print sample names, taxonomy ranks, and metadata fields — with explanations
cat("🔹 These are the metadata variables:\n")
print(sample_variables(ps_18s))
```

### Read depth per sample
```{r}
read_depth <- data.frame(
Sample = sample_names(ps_18s),
Reads = sample_sums(ps_18s),
check.names = FALSE
)

ggplot(read_depth, aes(x = Sample, y = Reads)) +
geom_col() +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
scale_x_discrete(
    labels = function(x) substr(x, 1, 6), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  ) +
labs(
title = "Sequencing depth per sample",
x = "Sample",
y = "Number of non-chimeric reads"
)
```

# COI Dataset

## 📦 Build the phyloseq object (COI)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)

# ---- paths ----
datadir <- "data/COI"

# ---- load DADA2 outputs ----
seqtab.nochim <- readRDS(file.path(datadir, "seqtab_nochim.rds"))
taxa          <- readRDS(file.path(datadir, "taxa.rds"))

# Make sure taxonomy is a proper matrix and has consistent column names
taxa <- as.matrix(taxa)
target_cols <- c(
  "Kingdom","Subkingdom_1","Subkingdom_2","Phylum","Subphylum","Subphylum_1","Superclass",
  "Class","Subclass","Infraclass","Superorder","Order","Suborder_1","Suborder_2","Suborder_3",
  "Family","Subfamily","Genus","Subgenus","Species"
)
colnames(taxa) <- target_cols[seq_len(ncol(taxa))]

# Load metadata
metadat <- readRDS(file.path(datadir, "metadata_COI.rds")) 

# Create a Phyloseq object

rownames(seqtab.nochim) <- sub("_$", "", rownames(seqtab.nochim))

ps_coi <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = F), 
               sample_data(metadat), 
               tax_table(taxa))

saveRDS(ps_coi, "data/ps_coi.rds")
cat("🔹 ** A summary of the phyloseq object**\n")
ps_coi

# storing the ASV sequences and assigning them with a name and then renaming the taxa object
dna <- Biostrings::DNAStringSet(taxa_names(ps_coi))
names(dna) <- taxa_names(ps_coi)
ps_coi <- merge_phyloseq(ps_coi, dna)
taxa_names(ps_coi) <- paste0("ASV", seq(ntaxa(ps_coi)))

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Print sample names, taxonomy ranks, and metadata fields — with explanations
cat("🔹 **These are the metadata variables:**\n")
print(sample_variables(ps_coi))
```

### Read depth per sample
```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(ggplot2)
library(scales)

read_depth <- data.frame(
Sample = sample_names(ps_coi),
Reads = sample_sums(ps_coi),
check.names = FALSE
)

ggplot(read_depth, aes(x = Sample, y = Reads)) +
  geom_col() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
  scale_x_discrete(
    labels = function(x) substr(x, 1, 6), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  ) +
  scale_y_continuous(
    labels = label_number(big.mark = ",") # Full numbers with commas
  ) +
  labs(
  title = "Sequencing depth per sample",
  x = "Sample",
  y = "Number of non-chimeric reads"
  )
```

# CPR Dataset

## 📦 Build the phyloseq object (CPR data)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)

# ---- load data ----
counts <- read_csv("data/count_table_CPR.csv")
taxo <- read_csv("data/taxonomy_table_CPR.csv")
meta <- read_csv("data/metadata_CPR.csv")

taxtab <- as.matrix(taxo)
row.names(taxtab) <- taxo[[8]]

# Create a Phyloseq object

ps_cpr <- phyloseq(otu_table(counts, taxa_are_rows = F), 
                   sample_data(meta), 
                   tax_table(taxtab))
saveRDS(ps_cpr, "data/ps_cpr.rds")
cat("🔹 ** A summary of the phyloseq object**\n")
print(ps_cpr)
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Print sample names, taxonomy ranks, and metadata fields — with explanations
cat("🔹 **These are the metadata variables:**\n")
print(sample_variables(ps_cpr))
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Print sample names, taxonomy ranks, and metadata fields — with explanations
cat("🔹 **These are the taxonomic levels:**\n")
print(rank_names(ps_cpr))
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Print sample names, taxonomy ranks, and metadata fields — with explanations
cat("🔹 **These are the metadata variables:**\n")
print(sample_variables(ps_cpr))
```

### Counts per sample
```{r, echo=FALSE, message=FALSE, warning=FALSE}
read_depth <- data.frame(
Sample = sample_names(ps_cpr),
Reads = sample_sums(ps_cpr),
check.names = FALSE
)

ggplot(read_depth, aes(x = Sample, y = Reads)) +
geom_col() +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for better readability
scale_x_discrete(
    labels = function(x) substr(x, 1, 6), # Show only the first 5 characters of each label
    breaks = function(x) x[seq(1, length(x), by = 4)] # Show every 4th label
  ) +
labs(
title = "CPR overall counts per sample",
x = "Sample",
y = "Number of counts"
)
```

# Merging datasets

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(phyloseq)
library(ggVennDiagram)

ps_18s <- readRDS("data/ps_18s.rds")
ps_coi <- readRDS("data/ps_coi.rds")
ps_cpr <- readRDS("data/ps_cpr.rds")

# 1) Make taxa IDs unique across markers
taxa_names(ps_18s) <- paste0("18S__", taxa_names(ps_18s))
taxa_names(ps_coi) <- paste0("COI__", taxa_names(ps_coi))

# 2) Add a marker column in the tax_table
add_marker <- function(ps, marker) {
  tt <- as(tax_table(ps), "matrix")
  tt <- cbind(tt, marker = marker)
  tax_table(ps) <- tax_table(tt)
  ps
}
ps_18s <- add_marker(ps_18s, "18S")
ps_coi <- add_marker(ps_coi, "COI")

# 3) Ensure tax_table columns match (pad missing cols with NA)
cols_all <- union(colnames(tax_table(ps_18s)), colnames(tax_table(ps_coi)))

pad_tax <- function(ps, cols_all) {
  tt <- as(tax_table(ps), "matrix")
  miss <- setdiff(cols_all, colnames(tt))
  if (length(miss) > 0) {
    tt <- cbind(tt, matrix(NA, nrow = nrow(tt), ncol = length(miss),
                          dimnames = list(rownames(tt), miss)))
  }
  tt <- tt[, cols_all, drop = FALSE]
  tax_table(ps) <- tax_table(tt)
  ps
}
ps_18s <- pad_tax(ps_18s, cols_all)
ps_coi <- pad_tax(ps_coi, cols_all)

# 4) Merge (stacks taxa/features)
ps_edna <- merge_phyloseq(ps_18s, ps_coi)

ps_edna
table(tax_table(ps_edna)[, "marker"])

```

### compare presence/absence in eDNA vs CPR

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pa_transform <- function(ps) {
  ps_pa <- ps
  otu_table(ps_pa) <- otu_table((otu_table(ps) > 0) * 1, taxa_are_rows(ps))
  ps_pa
}

ps_edna_pa <- pa_transform(ps_edna)
ps_cpr_pa  <- pa_transform(ps_cpr)

# 2 get taxa present at a given taxonomic rank

get_taxa_at_rank <- function(ps, rank) {
  tt <- as.data.frame(tax_table(ps))
  
  # taxa present at least once
  present_taxa <- taxa_names(ps)[taxa_sums(ps) > 0]
  
  # extract taxonomic labels
  taxa <- tt[present_taxa, rank]
  
  taxa <- taxa[!is.na(taxa) & taxa != ""]
  unique(as.character(taxa))
}

# 3 Compare eDNA vs CPR across taxonomic levels
ranks_to_test <- intersect(rank_names(ps_edna), rank_names(ps_cpr))

comparison <- lapply(ranks_to_test, function(rank) {
  edna_taxa <- get_taxa_at_rank(ps_edna_pa, rank)
  cpr_taxa  <- get_taxa_at_rank(ps_cpr_pa,  rank)

  data.frame(
    rank = rank,
    n_edna = length(edna_taxa),
    n_cpr  = length(cpr_taxa),
    n_shared = length(intersect(edna_taxa, cpr_taxa)),
    stringsAsFactors = FALSE
  )
})

taxa_overlap_summary <- do.call(rbind, comparison)

taxa_overlap_summary$perc_shared_edna <- with(
  taxa_overlap_summary,
  100 * n_shared / n_edna
)

taxa_overlap_summary$perc_shared_cpr <- with(
  taxa_overlap_summary,
  100 * n_shared / n_cpr
)

# optional: round for nicer tables
taxa_overlap_summary$perc_shared_edna <- round(taxa_overlap_summary$perc_shared_edna, 1)
taxa_overlap_summary$perc_shared_cpr  <- round(taxa_overlap_summary$perc_shared_cpr, 1)

knitr::kable(
  taxa_overlap_summary,
  caption = "Taxonomic overlap between eDNA and CPR datasets (presence/absence).",
  digits = 1
) |>
  kableExtra::kable_styling(full_width = FALSE)

```

### Combined figures

#### zooplankton

```{r}
# ---- genera to plot ----
genera10 <- c(
  "Paracalanus","Clausocalanus","Nannocalanus","Centropages","Acartia",
  "Calanus","Temora","Metridia","Oithona","Oncaea"
)

# ---- palette + named colours for the 10 genera ----
my_colors <- c("#FF6666", "#8DD3C7", "#377EB8", "#4DAF4A",
               "#984EA3", "#A6D854", "#FFFF33", "#FF7F00",
               "#F781BF", "#999999")

genus_cols <- setNames(my_colors[1:length(genera10)], genera10)

# ---- helper: phyloseq -> long df with longitude + Genus + PA ----
ps_to_long_df <- function(ps_obj, approach_name, genera_keep, lon_var = NULL, tax_rank = "Genus") {
  stopifnot(inherits(ps_obj, "phyloseq"))

  otu <- as(otu_table(ps_obj), "matrix")
  if (!taxa_are_rows(ps_obj)) otu <- t(otu)
  otu <- (otu > 0) * 1

  tax <- as(tax_table(ps_obj), "matrix") %>% as.data.frame(check.names = FALSE)
  tax$.taxa_id <- rownames(tax)

  sam <- as(sample_data(ps_obj), "data.frame") %>%
    rownames_to_column("SampleID")
  names(sam) <- trimws(names(sam))

  if (is.null(lon_var)) {
    lon_candidates <- names(sam)[grepl("^lon(gitude)?$", tolower(names(sam)))]
    if (length(lon_candidates) == 0) lon_candidates <- names(sam)[grepl("lon|long", tolower(names(sam)))]
    if (length(lon_candidates) == 0) {
      stop(
        "Could not find a longitude column in sample_data(ps_obj).\n",
        "Available columns:\n  ", paste(names(sam), collapse = ", ")
      )
    }
    lon_var <- lon_candidates[1]
  } else {
    if (!lon_var %in% names(sam)) {
      stop(sprintf(
        "Longitude column '%s' not found. Available columns:\n  %s",
        lon_var, paste(names(sam), collapse = ", ")
      ))
    }
  }

  if (!tax_rank %in% names(tax)) {
    stop(sprintf("Tax table column '%s' not found. Available tax ranks:\n  %s",
                 tax_rank, paste(names(tax), collapse = ", ")))
  }

  otu_long <- as.data.frame(otu) %>%
    rownames_to_column(".taxa_id") %>%
    pivot_longer(cols = -".taxa_id", names_to = "SampleID", values_to = "pa")

  otu_long %>%
    left_join(tax, by = ".taxa_id") %>%
    left_join(sam, by = "SampleID") %>%
    mutate(
      Genus = as.character(.data[[tax_rank]]),
      approach = approach_name,
      longitude = as.numeric(.data[[lon_var]])
    ) %>%
    filter(!is.na(Genus), Genus %in% genera_keep) %>%
    select(approach, SampleID, longitude, Genus, pa)
}

# ---- transect binning settings ----
lon_min <- -16.149
lon_max <- -1.246
n_bins  <- 10
lon_breaks <- seq(lon_min, lon_max, length.out = n_bins + 1)

# ---- build binned presence/absence by genus (any detection in bin = present) ----
df_binned <- bind_rows(
  ps_to_long_df(ps_cpr_pa,  "CPR",  genera10, lon_var = NULL, tax_rank = "Genus"),
  ps_to_long_df(ps_edna_pa, "eDNA", genera10, lon_var = NULL, tax_rank = "Genus")
) %>%
  filter(!is.na(longitude), longitude >= lon_min, longitude <= lon_max) %>%
  mutate(lon_bin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = TRUE)) %>%
  group_by(approach, lon_bin, Genus) %>%
  summarise(pa = as.integer(any(pa == 1)), .groups = "drop") %>%
  mutate(
    approach = factor(approach, levels = c("CPR", "eDNA"))
  )

# ---- rank genera by prevalence (most present across bins+approaches = top) ----
genus_rank <- df_binned %>%
  group_by(Genus) %>%
  summarise(prevalence = sum(pa == 1, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(prevalence), Genus)

genus_levels <- genus_rank$Genus

# apply ordering + fill mapping
df_binned <- df_binned %>%
  mutate(
    Genus = factor(Genus, levels = genus_levels),
    fill_genus = ifelse(pa == 1, as.character(Genus), NA)
  )

# reorder colours to match ranked genera
genus_cols_ranked <- genus_cols[genus_levels]

# ---- x-axis labels: bin midpoints ----
bin_mids <- (lon_breaks[-1] + lon_breaks[-length(lon_breaks)]) / 2
mid_labels <- sprintf("%.2f", bin_mids)
names(mid_labels) <- levels(df_binned$lon_bin)

# ---- plot (match phytoplankton style) ----
p <- ggplot(df_binned, aes(x = lon_bin, y = Genus, fill = fill_genus)) +
  geom_tile(width = 0.95, height = 0.9, colour = "grey80") +
  facet_wrap(~ approach, ncol = 1) +
  scale_fill_manual(
    values = genus_cols_ranked,
    na.value = "white",
    breaks = genus_levels,
    limits = genus_levels,
    drop = FALSE,
    name = "Genus"
  ) +
  scale_x_discrete(labels = mid_labels) +
  scale_y_discrete(limits = rev(genus_levels)) +  # most prevalent at TOP
  labs(x = "Longitude section (bin midpoint)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank()
  )

print(p)
```

#### Phytoplankton

```{r}

# ---- genera to plot ----
families9 <- c(
  "Rhizosoleniaceae",
  "Chaetoceraceae",
  "Ceratiaceae",
  "Coccolithaceae",
  "Dictyochaceae",
  "Leptocylindraceae",
  "Noelaerhabdaceae",
  "Rhabdosphaeraceae",
  "Prorocentraceae"
)

# ---- your palette + named colours for the 10 genera ----
my_colors <- c("#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#FFD92F",
               "#E5C494", "#B3B3B3", "#A65628","#FFFFB3", "#BEBADA",
               "#FB8072", "#80B1D3", "#FDB462")

fam_cols <- setNames(my_colors[1:length(families9)], families9)

# ---- helper: phyloseq -> long df with longitude + Family + PA ----
ps_to_long_df <- function(ps_obj, approach_name, families_keep, lon_var = NULL, tax_rank = "Family") {
  stopifnot(inherits(ps_obj, "phyloseq"))

  # OTU (taxa x samples) presence/absence
  otu <- as(otu_table(ps_obj), "matrix")
  if (!taxa_are_rows(ps_obj)) otu <- t(otu)
  otu <- (otu > 0) * 1

  # taxonomy
  tax <- as(tax_table(ps_obj), "matrix") %>% as.data.frame(check.names = FALSE)
  tax$.taxa_id <- rownames(tax)

  # sample data
  sam <- as(sample_data(ps_obj), "data.frame") %>%
    rownames_to_column("SampleID")
  names(sam) <- trimws(names(sam))

  # detect longitude column if not provided
  if (is.null(lon_var)) {
    lon_candidates <- names(sam)[grepl("^lon(gitude)?$", tolower(names(sam)))]
    if (length(lon_candidates) == 0) lon_candidates <- names(sam)[grepl("lon|long", tolower(names(sam)))]
    if (length(lon_candidates) == 0) {
      stop(
        "Could not find a longitude column in sample_data(ps_obj).\n",
        "Available columns:\n  ", paste(names(sam), collapse = ", ")
      )
    }
    lon_var <- lon_candidates[1]
  } else {
    if (!lon_var %in% names(sam)) {
      stop(sprintf(
        "Longitude column '%s' not found. Available columns:\n  %s",
        lon_var, paste(names(sam), collapse = ", ")
      ))
    }
  }

  if (!tax_rank %in% names(tax)) {
    stop(sprintf("Tax table column '%s' not found. Available tax ranks:\n  %s",
                 tax_rank, paste(names(tax), collapse = ", ")))
  }

  otu_long <- as.data.frame(otu) %>%
    rownames_to_column(".taxa_id") %>%
    pivot_longer(cols = -".taxa_id", names_to = "SampleID", values_to = "pa")

  otu_long %>%
    left_join(tax, by = ".taxa_id") %>%
    left_join(sam, by = "SampleID") %>%
    mutate(
      Family = as.character(.data[[tax_rank]]),
      approach = approach_name,
      longitude = as.numeric(.data[[lon_var]])
    ) %>%
    filter(!is.na(Family), Family %in% families_keep) %>%
    select(approach, SampleID, longitude, Family, pa)
}

# ---- transect binning settings ----
lon_min <- -16.149
lon_max <- -1.246
n_bins  <- 10
lon_breaks <- seq(lon_min, lon_max, length.out = n_bins + 1)

# ---- build binned presence/absence by family (any detection in bin = present) ----
# ---- build binned presence/absence by family (any detection in bin = present) ----
df_binned <- bind_rows(
  ps_to_long_df(ps_cpr_pa,  "CPR",  families9, lon_var = NULL, tax_rank = "Family"),
  ps_to_long_df(ps_edna_pa, "eDNA", families9, lon_var = NULL, tax_rank = "Family")
) %>%
  filter(!is.na(longitude), longitude >= lon_min, longitude <= lon_max) %>%
  mutate(lon_bin = cut(longitude, breaks = lon_breaks, include.lowest = TRUE, right = TRUE)) %>%
  group_by(approach, lon_bin, Family) %>%
  summarise(pa = as.integer(any(pa == 1)), .groups = "drop") %>%
  mutate(
    approach = factor(approach, levels = c("CPR", "eDNA"))
  )

# ---- rank families (top = most prevalent across bins and approaches) ----
fam_rank <- df_binned %>%
  group_by(Family) %>%
  summarise(
    # prevalence across ALL bins/approaches
    prevalence = sum(pa == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(prevalence), Family)

fam_levels <- fam_rank$Family  # most prevalent first

# ---- apply ordering + fill mapping ----
df_binned <- df_binned %>%
  mutate(
    Family = factor(Family, levels = fam_levels),  # rows/legend order control
    fill_family = ifelse(pa == 1, as.character(Family), NA)
  )

# reorder colours to match the ranked families
fam_cols_ranked <- fam_cols[fam_levels]

# ---- x-axis labels: bin midpoints ----
bin_mids <- (lon_breaks[-1] + lon_breaks[-length(lon_breaks)]) / 2
mid_labels <- sprintf("%.2f", bin_mids)
names(mid_labels) <- levels(df_binned$lon_bin)

# ---- plot ----
p <- ggplot(df_binned, aes(x = lon_bin, y = Family, fill = fill_family)) +
  geom_tile(width = 0.95, height = 0.9, colour = "grey80") +
  facet_wrap(~ approach, ncol = 1) +
  scale_fill_manual(
    values = fam_cols_ranked,
    na.value = "white",
    breaks = fam_levels,
    limits = fam_levels,
    drop = FALSE,
    name = "Family"
  ) +
  scale_x_discrete(labels = mid_labels) +
  scale_y_discrete(limits = rev(fam_levels)) +
  labs(x = "Longitude section (bin midpoint)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),

    # 🔹 Remove y-axis completely
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank()
  )

print(p)
```

#### Prevalence scatterplot

```{r, fig.width=12, fig.height=8}
library(tidyverse)
library(phyloseq)
library(gridExtra)
library(patchwork)

# -----------------------------
# 1. Shared genera
# -----------------------------
shared_genera <- intersect(
  get_taxa_at_rank(ps_edna_pa, "Genus"),
  get_taxa_at_rank(ps_cpr_pa, "Genus")
)

# -----------------------------
# 2. Function to calculate genus prevalence
# -----------------------------
get_prev <- function(ps, method, genera_keep) {
  
  otu <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) otu <- t(otu)
  
  tax <- as(tax_table(ps), "matrix") %>%
    as.data.frame(check.names = FALSE) %>%
    rownames_to_column("taxa_id")
  
  as.data.frame(otu > 0) %>%
    rownames_to_column("taxa_id") %>%
    pivot_longer(
      -taxa_id,
      names_to = "SampleID",
      values_to = "present"
    ) %>%
    left_join(tax, by = "taxa_id") %>%
    mutate(Genus = as.character(Genus)) %>%
    filter(!is.na(Genus), Genus %in% genera_keep) %>%
    group_by(SampleID, Genus) %>%
    summarise(present = any(present), .groups = "drop") %>%
    complete(
      SampleID = sample_names(ps),
      Genus = genera_keep,
      fill = list(present = FALSE)
    ) %>%
    group_by(Genus) %>%
    summarise(
      prevalence = mean(present) * 100,
      .groups = "drop"
    ) %>%
    mutate(method = method)
}

# -----------------------------
# 3. Broad taxonomic group
# -----------------------------
get_group <- function(ps, genera_keep) {
  
  as(tax_table(ps), "matrix") %>%
    as.data.frame(check.names = FALSE) %>%
    mutate(across(everything(), as.character)) %>%
    filter(!is.na(Genus), Genus %in% genera_keep) %>%
    distinct(Genus, Phylum, Class) %>%
    mutate(
      broad_group = case_when(
        Phylum %in% c("Arthropoda", "Cnidaria", "Ctenophora", "Mollusca",
                      "Annelida", "Chaetognatha", "Chordata", "Rotifera",
                      "Ciliophora") ~ "Zooplankton",
        
        Phylum %in% c("Bacillariophyta", "Dinoflagellata", "Haptophyta",
                      "Ochrophyta", "Chlorophyta", "Cryptophyta",
                      "Cyanobacteria") |
          Class %in% c("Mediophyceae", "Bacillariophyceae",
                       "Coscinodiscophyceae", "Dinophyceae",
                       "Prymnesiophyceae", "Dictyochophyceae",
                       "Chlorophyceae", "Cryptophyceae") ~ "Phytoplankton",
        
        TRUE ~ "Other/ambiguous"
      )
    ) %>%
    group_by(Genus) %>%
    summarise(
      broad_group = case_when(
        any(broad_group == "Zooplankton") ~ "Zooplankton",
        any(broad_group == "Phytoplankton") ~ "Phytoplankton",
        TRUE ~ "Other/ambiguous"
      ),
      .groups = "drop"
    )
}

# -----------------------------
# 4. Combine prevalence and taxonomy
# -----------------------------
genus_prev_df <- bind_rows(
  get_prev(ps_cpr_pa,  "CPR",  shared_genera),
  get_prev(ps_edna_pa, "eDNA", shared_genera)
) %>%
  pivot_wider(
    names_from = method,
    values_from = prevalence
  )

genus_groups <- bind_rows(
  get_group(ps_cpr_pa, shared_genera),
  get_group(ps_edna_pa, shared_genera)
) %>%
  group_by(Genus) %>%
  summarise(
    broad_group = case_when(
      any(broad_group == "Zooplankton") ~ "Zooplankton",
      any(broad_group == "Phytoplankton") ~ "Phytoplankton",
      TRUE ~ "Other/ambiguous"
    ),
    .groups = "drop"
  )

plot_df <- genus_prev_df %>%
  left_join(genus_groups, by = "Genus") %>%
  arrange(desc(eDNA - CPR)) %>%
  mutate(
    taxon_id = row_number(),
    label = paste0(taxon_id, ". ", Genus)
  )

# -----------------------------
# 5. Spearman correlation
# -----------------------------
cor_test <- cor.test(
  plot_df$CPR,
  plot_df$eDNA,
  method = "spearman"
)

rho_label <- paste0(
  "n = ", nrow(plot_df),
  "\nSpearman \u03c1 = ", round(unname(cor_test$estimate), 2),
  "\np = ", signif(cor_test$p.value, 2)
)

# -----------------------------
# 6. Plot with numbers instead of points
# -----------------------------
library(ggplot2)

p_fig3 <- ggplot(plot_df, aes(x = CPR, y = eDNA)) +
    annotate(
    "polygon",
    x = c(0, 100, 0),
    y = c(0, 100, 100),
    fill = "#DCEAF7",
    alpha = 0.4
  ) +

  annotate(
    "polygon",
    x = c(0, 100, 100),
    y = c(0, 0, 100),
    fill = "#EEEEEE",
    alpha = 0.6
  ) +
    annotate(
    "text",
    x = 55,
    y = 100,
    label = "Higher prevalence\nin eDNA",
    colour = "#3B6FB6",
    fontface = "italic",
    hjust = 0,
    vjust = 1,
    size = 4,
    alpha = 0.8
  ) +
  
  annotate(
    "text",
    x = 73,
    y = 60,
    label = "Higher prevalence\nin CPR",
    colour = "grey40",
    fontface = "italic",
    hjust = 0,
    size = 4
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    colour = "grey55",
    linewidth = 0.5
  ) +

  geom_point(
    aes(fill = broad_group),
    shape = 21,
    size = 6.5,
    colour = "black",
    stroke = 0.45,
    alpha = 0.95
  ) +

  geom_text(
    aes(label = taxon_id),
    colour = "white",
    fontface = "bold",
    size = 3.2
  ) +
  scale_fill_manual(
    values = c(
      "Zooplankton"   = "#D95F02",  # orange-red
      "Phytoplankton" = "#1B7837"   # deep green
    )
  ) +

  scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 25),
    expand = expansion(mult = c(0.02, 0.03))
  ) +

  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 25),
    expand = expansion(mult = c(0.02, 0.03))
  ) +

  coord_equal(clip = "off") +

  labs(
    x = "Prevalence in CPR samples (%)",
    y = "Prevalence in eDNA samples (%)",
    title = "Comparison of genus prevalence between CPR and eDNA"
  ) +

  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(colour = "black"),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 11),
    legend.key = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.4),
    plot.margin = margin(8, 12, 8, 8)
  )

# -----------------------------
# 7. Side list of numbered taxa
# -----------------------------

taxa_table <- plot_df %>%
  arrange(taxon_id) %>%
  select(taxon_id, Genus)

n <- nrow(taxa_table)

tab1 <- taxa_table[1:ceiling(n/2), ]
tab2 <- taxa_table[(ceiling(n/2)+1):n, ]

g1 <- tableGrob(tab1, rows = NULL)
g2 <- tableGrob(tab2, rows = NULL)

table_plot <- arrangeGrob(
  g1, g2,
  ncol = 2
)

p_final <-
  p_fig3 +
  wrap_elements(table_plot) +
  plot_layout(widths = c(2.6, 1.4))

p_final
```

### NMDS

#### CPR

```{r}
library(phyloseq)
library(vegan)
library(tidyverse)

sample_data(ps_cpr)$Water_masses <- ifelse(sample_data(ps_cpr)$Longitude < -10,
                                       "Open Ocean", "Shelf")

# Make sure Water_masses is a factor (optional, but useful for consistent plotting)
sample_data(ps_cpr)$Water_masses <- factor(sample_data(ps_cpr)$Water_masses, 
                                       levels = c("Open Ocean", "Shelf"))

otu_cpr <- otu_table(ps_cpr)
tax_cpr <- tax_table(ps_cpr)
sam_cpr <- sample_data(ps_cpr) 

# force taxonomy order to match taxa order 
tax_cpr <- tax_cpr[taxa_names(ps_cpr), ] 

# rebuild object
ps_cpr_fixed <- phyloseq(otu_cpr, tax_cpr, sam_cpr)

# ---- extract CPR community matrix ----
mat_cpr <- as(otu_table(ps_cpr_fixed), "matrix")

# samples should be rows, taxa should be columns
if (taxa_are_rows(ps_cpr_fixed)) {
  mat_cpr <- t(mat_cpr)
}

# ---- convert to presence/absence ----
mat_cpr_pa <- (mat_cpr > 0) * 1

# remove taxa absent from all samples
mat_cpr_pa <- mat_cpr_pa[, colSums(mat_cpr_pa) > 0, drop = FALSE]

# remove samples with no taxa detected, if any
mat_cpr_pa <- mat_cpr_pa[rowSums(mat_cpr_pa) > 0, , drop = FALSE]

# ---- NMDS using Jaccard distance ----
set.seed(123)

nmds_cpr <- metaMDS(
  mat_cpr_pa,
  distance = "jaccard",
  binary = TRUE,
  k = 2,
  trymax = 200,
  autotransform = FALSE,
  trace = FALSE
)

# ---- extract NMDS scores ----
scores_cpr <- as.data.frame(scores(nmds_cpr, display = "sites")) %>%
  rownames_to_column("SampleID")

# ---- extract metadata ----
meta_cpr <- as(sample_data(ps_cpr_fixed), "data.frame") %>%
  rownames_to_column("SampleID")

# keep only samples used in NMDS
meta_cpr <- meta_cpr %>%
  filter(SampleID %in% rownames(mat_cpr_pa))

plot_cpr <- scores_cpr %>%
  left_join(meta_cpr, by = "SampleID")

ggplot(
  plot_cpr,
  aes(
    x = NMDS1,
    y = NMDS2,
    colour = Water_masses
  )
) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(
    aes(group = Water_masses),
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    values = c(
      "Open Ocean" = "#D55E00",
      "Shelf" = "#0072B2"
    )
  ) +
  coord_equal(
    xlim = c(-1.3, 1.0),
    ylim = c(-1.3, 1.0)
  ) +
  theme_minimal(base_size = 12)
```

```{r}
adonis2(
  mat_cpr_pa ~ Water_masses,
  data = meta_cpr,
  method = "jaccard",
  binary = TRUE,
  permutations = 999
)
```

```{r}
dist_cpr <- vegdist(
  mat_cpr_pa,
  method = "jaccard",
  binary = TRUE
)

disp_cpr <- betadisper(
  dist_cpr,
  meta_cpr$Water_masses
)

anova(disp_cpr)
permutest(disp_cpr)
```

#### eDNA NMDS at genus level

```{r}
library(phyloseq)
library(vegan)
library(tidyverse)

sample_data(ps_edna)$Water_masses <- ifelse(sample_data(ps_edna)$lon < -10,
                                       "Open Ocean", "Shelf")

# Make sure Water_masses is a factor (optional, but useful for consistent plotting)
sample_data(ps_edna)$Water_masses <- factor(sample_data(ps_edna)$Water_masses, 
                                       levels = c("Open Ocean", "Shelf"))

# -------------------------
# Aggregate to Genus level
# -------------------------

ps_edna_genus <- tax_glom(
  ps_edna,
  taxrank = "Genus",
  NArm = TRUE
)

# -------------------------
# Extract matrix
# -------------------------

mat_genus <- as(otu_table(ps_edna_genus), "matrix")

if (taxa_are_rows(ps_edna_genus)) {
  mat_genus <- t(mat_genus)
}

# -------------------------
# Presence / absence
# -------------------------

mat_genus_pa <- (mat_genus > 0) * 1

# Remove empty taxa
mat_genus_pa <- mat_genus_pa[
  ,
  colSums(mat_genus_pa) > 0,
  drop = FALSE
]

# Remove empty samples
mat_genus_pa <- mat_genus_pa[
  rowSums(mat_genus_pa) > 0,
  ,
  drop = FALSE
]

dim(mat_genus_pa)

# -------------------------
# NMDS
# -------------------------

set.seed(123)

nmds_edna_genus <- metaMDS(
  mat_genus_pa,
  distance = "jaccard",
  binary = TRUE,
  k = 2,
  trymax = 200,
  autotransform = FALSE,
  trace = FALSE
)

nmds_edna_genus$stress

scores_edna_genus <- as.data.frame(
  scores(nmds_edna_genus, display = "sites")
) %>%
  rownames_to_column("SampleID")

meta_edna <- as(sample_data(ps_edna), "data.frame") %>%
  rownames_to_column("SampleID")

plot_edna_genus <- left_join(
  scores_edna_genus,
  meta_edna,
  by = "SampleID"
)

p_edna_genus <- ggplot(
  plot_edna_genus,
  aes(
    NMDS1,
    NMDS2,
    colour = Water_masses
  )
) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(
    aes(group = Water_masses),
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  scale_colour_manual(
    values = c(
      "Open Ocean" = "#D55E00",
      "Shelf" = "#0072B2"
    )
  ) +
  coord_equal(
    xlim = c(-1.3, 1.0),
    ylim = c(-1.3, 1.0)
  ) +
  labs(
    title = "eDNA (Metabarcoding)",
    colour = NULL
  ) +
  theme_minimal(base_size = 12)

p_edna_genus
```

```{r}
adonis2(
  mat_genus_pa ~ Water_masses,
  data = meta_edna,
  method = "jaccard",
  binary = TRUE,
  permutations = 999
)
```

```{r}
dist_genus <- vegdist(
  mat_genus_pa,
  method = "jaccard",
  binary = TRUE
)

disp_genus <- betadisper(
  dist_genus,
  meta_edna$Water_masses
)

anova(disp_genus)
permutest(disp_genus)
```

### Unique to CPR vs unique to eDNA

```{r genus-overlap-counts, cache=FALSE}
library(tidyverse)
library(phyloseq)

# -----------------------------
# 1. Helper: extract genera present in a phyloseq object
# -----------------------------

get_present_genera <- function(ps) {

  tax <- as(tax_table(ps), "matrix") %>%
    as.data.frame(check.names = FALSE)

  present_taxa <- taxa_names(ps)[taxa_sums(ps) > 0]

  tax[present_taxa, "Genus"] %>%
    as.character() %>%
    purrr::discard(~ is.na(.x) |
              .x == "" |
              .x == "N/A" |
              .x == "NA" |
              .x == "Unknown") %>%
    unique()
}

edna_genera <- get_present_genera(ps_edna)
cpr_genera  <- get_present_genera(ps_cpr)

shared_genera   <- intersect(edna_genera, cpr_genera)
cpr_only_genera <- setdiff(cpr_genera, edna_genera)
edna_only_genera <- setdiff(edna_genera, cpr_genera)

genus_overlap_counts <- tibble(
  category = c("Shared", "CPR only", "eDNA only"),
  n_genera = c(
    length(shared_genera),
    length(cpr_only_genera),
    length(edna_only_genera)
  )
)

genus_overlap_counts
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# -----------------------------
# 3. Helper: calculate genus prevalence within one method
# -----------------------------

calculate_genus_prevalence <- function(ps, genera_keep) {

  otu <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) otu <- t(otu)

  otu <- (otu > 0) * 1

  tax <- as(tax_table(ps), "matrix") %>%
    as.data.frame(check.names = FALSE) %>%
    rownames_to_column(".taxa_id")

  sample_ids <- sample_names(ps)

  as.data.frame(otu) %>%
    rownames_to_column(".taxa_id") %>%
    pivot_longer(
      cols = -".taxa_id",
      names_to = "SampleID",
      values_to = "pa"
    ) %>%
    left_join(tax, by = ".taxa_id") %>%
    mutate(
      Genus = as.character(Genus)
    ) %>%
    filter(!is.na(Genus), Genus %in% genera_keep) %>%
    group_by(SampleID, Genus) %>%
    summarise(
      pa = as.integer(any(pa == 1)),
      .groups = "drop"
    ) %>%
    complete(
      SampleID = sample_ids,
      Genus = genera_keep,
      fill = list(pa = 0)
    ) %>%
    group_by(Genus) %>%
    summarise(
      n_present = sum(pa == 1),
      n_samples = n_distinct(SampleID),
      prevalence = 100 * n_present / n_samples,
      .groups = "drop"
    ) %>%
    arrange(desc(prevalence))
}

# -----------------------------
# 4. Most prevalent CPR-only genera
# -----------------------------

cpr_only_prev <- calculate_genus_prevalence(
  ps = ps_cpr,
  genera_keep = cpr_only_genera
) %>%
  mutate(category = "CPR only")

cpr_only_prev %>%
  arrange(desc(prevalence)) %>%
  slice_head(n = 20)

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# -----------------------------
# 5. Most prevalent eDNA-only genera
# -----------------------------

edna_only_prev <- calculate_genus_prevalence(
  ps = ps_edna,
  genera_keep = edna_only_genera
) %>%
  mutate(category = "eDNA only")

edna_only_prev %>%
  arrange(desc(prevalence)) %>%
  slice_head(n = 20)
```

### Investigating the top 20 most prevalent eDNA-only taxa

```{r}
tax_edna <- as.data.frame(tax_table(ps_edna))

edna_only_taxonomy <- tax_edna %>%
  filter(Genus %in% edna_only_genera) %>%
  select(
    Phylum,
    Class,
    Order,
    Family,
    Genus
  ) %>%
  distinct()

rownames(edna_only_taxonomy) <- NULL

edna_only_summary <- edna_only_prev %>%
  left_join(
    edna_only_taxonomy,
    by = "Genus"
  ) %>%
  arrange(desc(prevalence))

head(edna_only_summary, 30)
```

```{r}

class_edna_summary <- edna_only_summary %>% group_by(Class) %>% summarise( n_genera = n_distinct(Genus), mean_prevalence = mean(prevalence), .groups = "drop" ) %>% arrange(desc(n_genera))

class_plot_df <- class_edna_summary %>%
  filter(mean_prevalence > 10, n_genera >= 5) %>%
  arrange(desc(n_genera)) %>%
  slice_head(n = 15) %>%   # adjust if you want more/fewer classes
  mutate(
    Class = forcats::fct_reorder(Class, n_genera)
  )

p_edna_only_classes <- ggplot(
  class_plot_df,
  aes(
    x = n_genera,
    y = Class,
    fill = mean_prevalence
  )
) +
  geom_col(width = 0.75) +
  geom_text(
    aes(label = n_genera),
    hjust = -0.25,
    size = 3.5
  ) +
  scale_fill_gradient(
    low = "grey85",
    high = "black",
    name = "Mean\nprevalence (%)"
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    x = "Number of genera detected only by eDNA",
    y = NULL,
    title = "Taxonomic classes containing genera uniquely detected by eDNA"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_edna_only_classes
```